Hypoxia-inducible factor promotes cysteine homeostasis in Caenorhabditis elegans

Version 1

Abstract

Abstract

The amino acid cysteine is critical for many aspects of life, yet excess cysteine is toxic. Therefore, animals require pathways to maintain cysteine homeostasis. In mammals, high cysteine activates cysteine dioxygenase, a key enzyme in cysteine catabolism. The mechanism by which cysteine dioxygenase is regulated remains largely unknown. We discovered that C. elegans cysteine dioxygenase (cdo-1) is transcriptionally activated by high cysteine and the hypoxia inducible transcription factor (hif-1). hif-1- dependent activation of cdo-1 occurs downstream of an H2S-sensing pathway that includes rhy-1, cysl-1, and egl-9. cdo-1 transcription is primarily activated in the hypodermis where it is sufficient to drive sulfur amino acid metabolism. EGL-9 and HIF-1 are core members of the cellular hypoxia response. However, we demonstrate that the mechanism of HIF-1-mediated induction of cdo-1 functions largely independent of EGL-9 prolyl hydroxylation and the von Hippel-Lindau E3 ubiquitin ligase; classical hypoxia signaling pathway components. We propose that the intersection of hif-1 and cdo-1 reveals a negative feedback loop for maintaining cysteine homeostasis. High cysteine stimulates the production of an H2S signal. H2S then activates the rhy-1/cysl-1/egl-9 signaling pathway, increasing HIF-1-mediated transcription of cdo-1, promoting degradation of cysteine via CDO-1.

Introduction

Cysteine is a sulfur-containing amino acid that mediates many oxidation/reduction reactions of proteins, is a substrate in the production of iron-sulfur clusters and hydrogen sulfide (H2S), a volatile signaling molecule, and a precursor to the antioxidant glutathione (1–3). Cysteine residues in close proximity in the primary or folded protein sequence are oxidized in the endoplasmic reticulum to form intra- and interprotein disulfide linkages, most commonly in secreted proteins which mediate intercellular signaling and defense (4–6). In many enzymes, the reactivity of the cysteine sulfur is key for the coordination of metals such as zinc or iron, which support protein structure and catalytic activity (7–9). In contrast to its essential function, excess cysteine is also toxic. High cysteine impairs mitochondrial respiration by disrupting iron homeostasis (10), acts as a neural excitotoxin (11), and promotes the formation of toxic levels of hydrogen sulfide gas (2, 12, 13). Given this balance between essential and toxic, cysteine homeostasis is key for the health of cells and organisms.

Cysteine is oxidized to cysteinesulfinate by cysteine dioxygenase (CDO-1 in C. elegans, CDO1 in mammals) (Fig. 1A) (14). CDO1-mediated oxidation is the primary pathway of cysteine catabolism when sulfur amino acid (methionine or cysteine) availability is normal or high (15). As a critical player in cysteine homeostasis, CDO1 is a highly regulated enzyme: the activity and abundance of CDO1 increase dramatically in cells and animals fed excess cysteine and methionine (16). CDO1 activation is governed both transcriptionally and post-translationally via the 26S proteasome (17–21). However, the molecular players that sense high cysteine and promote CDO1 activation remain incompletely defined.

egl-9 and rhy-1 inhibit cdo-1 transcription

A) Simplified pathway for sulfur amino acid metabolism beginning with methionine. We highlight the roles of cystathionase (CTH-2), cysteine dioxygenase (CDO-1), and the Moco-requiring sulfite oxidase enzyme (SUOX-1). Human homologs are displayed in purple. Abbreviations are as follows: Met (methionine), Ser (serine), Cth (cystathionine), Cys (cysteine), CSA (cysteinesulfinate), and αKB (α-ketobutyrate). B) Models of Pcdo-1::GFP (upper) and Pcdo-1::CDO-1::GFP (lower) transgenes used in this work are displayed. Boxes indicate exons, connecting lines indicate introns. The cdo-1 promoter is shown. C) egl-9a and rhy-1 gene structures are displayed. Boxes indicate exons and connecting lines are introns. Colored annotations indicate mutations generated or used in our work. Red; chemically-induced mutations that activated Pcdo-1::CDO-1::GFP. Blue; reference null alleles isolated independent of our work. Purple; CRISPR/Cas9-generated mutation that inactivates the prolyl hydroxylase domain of EGL-9. D) Expression of Pcdo-1::GFP transgene is displayed from wild-type, egl-9(sa307), and rhy-1(ok1402) C. elegans animals at the L4 stage of development. Scale bar is 250μm. E) Quantification of GFP expression displayed in Fig. 1D. Individual datapoints are shown (circles) as are the mean and standard deviation (red lines). n is 5 individuals per genotype. Data are normalized so that wild-type expression of Pcdo-1::GFP is 1 arbitrary unit (a.u.). ****, p<0.0001, ordinary one-way ANOVA with Dunnett’s post hoc analysis.

CDO1 dysfunction is also implicated in disease. CDO1 is a tumor suppressor whose activity is silenced in diverse cancers via promoter methylation, a potential biomarker of tumor grade and progression (22–24). Decreased CDO1 activity may support tumor cell growth by reducing reactive oxygen species and decreasing drug susceptibility (25–28).

Using the nematode C. elegans, we have shown that CDO-1 is a key player in the pathophysiology of 2 fatal inborn errors of metabolism; molybdenum cofactor deficiency (MoCD) and isolated sulfite oxidase deficiency (ISOD) (29, 30). Molybdenum cofactor (Moco) is an essential 520 Dalton prosthetic group synthesized from GTP by a conserved multistep biosynthetic pathway that is present in about 2/3 of bacterial genomes and nearly all eukaryotic genomes (31, 32). C. elegans can either retrieve Moco synthesized by the bacteria it consumes or can synthesize Moco de novo using its own Moco biosynthetic pathway (33). C. elegans strains carrying mutations in genes encoding Moco biosynthetic enzymes (moc) and feeding on wild-type E. coli develop normally. Yet, these same moc-mutant C. elegans fed Moco-deficient E. coli are completely inviable. A genetic selection for mutations that suppress this inviability identified mutations in cth-2 (cystathionase) and cdo-1 (cysteine dioxygenase), enzymes in the cysteine biosynthetic and degradation pathway. Using the dipeptide cystathionine as a substrate, CTH-2 produces α-ketobutyrate and cysteine. Cysteine is further oxidized by CDO-1, generating highly toxic sulfites. These sulfites are normally oxidized to more benign sulfate by Moco-requiring sulfite oxidase, SUOX-1, an essential enzyme in C. elegans and humans (Fig. 1A) (30, 33). Therefore, loss of cdo-1 or cth-2 suppresses the lethality caused by Moco deficiency in C. elegans by not producing sulfites which are toxic when SUOX-1 is inactive due to Moco deficiency (33–35). Thus, understanding the fundamental mechanisms that govern the levels and activity of CDO-1 is critical to generating new therapeutic hypotheses to treat these diseases.

To define genes that regulate cdo-1 levels and activity, we performed an unbiased genetic screen in the nematode C. elegans to identify mutations that increase the expression or abundance of CDO-1. We identified loss-of-function mutations in 2 genes, egl-9 and rhy-1, that dramatically increase expression of a Pcdo-1::CDO-1::GFP reporter transgene. These enzymes act in the hypoxia and H2S-sensing pathway, and we demonstrate that the conserved hypoxia-inducible transcription factor (HIF-1) activates cdo-1 transcription in this pathway (36–38). We further show that high cysteine promotes cdo-1 transcription, and that hif-1 and cysl-1 (another component of the H2S-sensing pathway) are required for viability under high cysteine conditions. We demonstrate that transcriptional activation of cdo-1 via HIF-1 promotes CDO-1 activity and establish the C. elegans hypodermis as a key tissue of CDO-1 activation and function. Unexpectedly, we find that cdo-1 regulation is governed by a HIF-1 pathway largely independent of EGL-9 prolyl hydroxylase activity and von Hippel-Lindau (VHL-1), the canonical oxygen-sensing pathway (39, 40). These data establish a new connection between the HIF-1/H2S-sensing pathway and sulfur amino acid catabolism governed by CDO-1.

Results

egl-9 and rhy-1 negatively regulate cdo-1 transcription

To identify regulators of CDO-1 levels, we engineered a transgene expressing a C-terminal green fluorescent protein (GFP) tagged version of full-length CDO-1 protein driven by the native cdo-1 promoter (Pcdo-1::CDO-1::GFP, Fig. 1B). Transgenic animals were generated by integrating the Pcdo-1::CDO-1::GFP construct into the C. elegans genome (41). This transgene was functional and rescued a cdo-1 loss of function mutation (Fig. S1). The Pcdo-1::CDO-1::GFP transgene reverses the suppression of Moco-deficient lethality caused by cdo-1 loss of function (Fig. S1). The reanimation of Moco-deficient lethality by the transgene depends on CDO-1 enzymatic activity because a transgene expressing an active-site mutant Pcdo-1::CDO-1[C85Y]::GFP does not rescue the cdo-1 mutant phenotype (Fig. S1) (33, 42). These data demonstrate that the Pcdo-1::CDO-1::GFP reporter transgene is functional and suggests its expression pattern reflects endogenous protein expression, localization, and levels.

We performed a mutagenesis screen to identify genes that control the expression or accumulation of CDO-1 protein. Specifically, we performed an EMS chemical mutagenesis of C. elegans and screened for mutations that caused increased GFP accumulation by the Pcdo-1::CDO-1::GFP reporter transgene. New mutants were isolated two generations post mutagenesis to allow novel mutations to become homozygous (43). Using whole-genome sequencing, we determined that 2 independently isolated mutants carried distinct mutations in egl-9 and four other independently isolated mutants had unique mutations in rhy-1 (Fig. 1C). The presence of multiple independent alleles suggests these mutations in egl-9 or rhy-1 are causative for the increased Pcdo-1::CDO-1::GFP expression or accumulation phenotype. egl-9 encodes the oxygen-sensing prolyl hydroxylase and orthologue of mammalian EGLN1. rhy-1 encodes the regulator of hypoxia inducible transcription factor, an enzyme with homology to membrane-bound O-acyltransferases (36, 39). The genetic screen produced nonsense alleles of both egl-9 and rhy-1 suggesting the increased Pcdo-1::CDO-1::GFP expression or accumulation phenotype is caused by loss of egl-9 or rhy-1 function (Fig. 1C). Inactivation of egl-9 or rhy-1 activates a transcriptional program mediated by the hypoxia inducible transcription factor, HIF-1 (36, 39). To determine if egl-9 or rhy-1 regulate cdo-1 transcription, we engineered a separate reporter construct where only GFP is transcribed by the cdo-1 promoter (Pcdo-1::GFP, Fig. 1B). This Pcdo-1::GFP transgene was introduced into strains with independently isolated egl-9(sa307) and rhy-1(ok1402) null reference alleles (36, 44). The egl-9(sa307) and rhy-1(ok1402) mutations dramatically induce GFP driven by the Pcdo-1::GFP transgene (Fig. 1D,E). The activation of cdo-1 transcription by independently isolated egl-9 or rhy-1 mutations demonstrates that the egl-9 and rhy-1 mutations isolated in our screen for the induction or accumulation of Pcdo-1::CDO-1::GFP are the causative genetic lesions. Furthermore, these data demonstrate that egl-9 and rhy-1 are necessary for the normal transcriptional repression of cdo-1.

HIF-1 directly activates cdo-1 transcription downstream of the rhy-1, cysl-1, egl-9 genetic pathway

rhy-1 and egl-9 act in a pathway that regulates the abundance and activity of the HIF-1 transcription factor (36, 38). The activity of rhy-1 is most upstream in the pathway and negatively regulates the activity of cysl-1, which encodes a cysteine synthase-like enzyme likely of algal origin (45). CYSL-1 directly binds to and inhibits EGL-9 in an H2S-modulated manner (38). EGL-9 uses molecular oxygen as well as an α-ketoglutarate cofactor to directly inhibit HIF-1 via prolyl hydroxylation, which recruits the VHL-1 ubiquitin ligase to ubiquitinate HIF-1, targeting it for degradation by the proteasome (46–48). Given that loss-of-function mutations in rhy-1 or egl-9 activate cdo-1 transcription, we tested if cdo-1 transcription is activated by HIF-1 as an output of this hypoxia/H2S-sensing pathway. We performed epistasis studies using null mutations that inhibit the activity of hif-1 (cysl-1(ok762) and hif-1(ia4)) or activate hif-1 (rhy-1(ok1402) and egl-9(sa307)). The induction of Pcdo-1::GFP by egl-9 inactivation was dependent upon the activity of hif-1, but not on the activity of cysl-1 (Fig. 2A,B). In contrast, induction of Pcdo-1::GFP by rhy-1 inactivation was dependent upon the activity of both hif-1 and cysl-1 (Fig. 2A,C). These results reveal a genetic pathway whereby rhy-1, cysl-1, and egl-9 function in a negative-regulatory cascade to control the activity of HIF-1 which transcriptionally activates cdo-1. Our epistasis studies of cdo-1 transcriptional regulation by HIF-1 align well with previous analyses of this genetic pathway in the context of transcription of cysl-2 (a paralog of cysl-1) and the “O2-ON response” (38).

cdo-1 transcription is activated by hif-1 downstream of rhy-1, cysl-1, and egl-9

A) Expression of the Pcdo-1::GFP transgene is displayed for wild-type, egl-9(sa307), egl-9(sa307) hif-1(ia4) double mutant, egl-9(sa307); cysl-1(ok762) double mutant, rhy-1(ok1402), rhy-1(ok1402); hif-1(ia4) double mutant, and rhy-1(ok1402); cysl-1(ok762) double mutant C. elegans animals at the L4 stage of development. Scale bar is 250μm. B,C) Quantification of the data displayed in Fig. 2A. Individual datapoints are shown (circles) as are the mean and standard deviation (red lines). n is 5 individuals per genotype. Data are normalized so that wild-type expression of Pcdo-1::GFP is 1 arbitrary unit (a.u.). *, p<0.05, ****, p<0.0001, ordinary one-way ANOVA with Dunnett’s post hoc analysis. Note, wild-type, egl-9(-), and rhy-1(-) images in panel A and quantification of Pcdo-1::GFP in panels B and C are identical to the data presented in Fig. 1D,E. They are re-displayed here to allow for clear comparisons to the double mutant strains of interest.

To demonstrate that HIF-1 activates transcription of endogenous cdo-1, we explored published RNA-sequencing data of wild-type, egl-9(-), and egl-9(-) hif-1(-) mutant animals (49). egl-9(-) mutant C. elegans display an 8-fold increase in cdo-1 mRNA compared to wild type. This induction was dependent on hif-1; a hif-1(-) mutation completely suppressed the induction of cdo-1 mRNA caused by an egl-9(-) mutation (49). These RNA-seq data confirm our findings using the Pcdo-1::GFP transcriptional reporter that HIF-1 is a transcriptional activator of cdo-1. ChIP-seq data of HIF-1 performed by the modERN project show that HIF-1 directly binds the cdo-1 promoter (peak from −1,165 to −714 base pairs 5’ to the cdo-1 ATG start codon) (50). This HIF-1 binding site contains 3 copies of the HIF-binding motif (5’-RCGTG-3’) (51). We conclude that cdo-1 is a downstream effector of HIF-1 and is likely a direct transcriptional target of HIF-1.

High cysteine promotes cdo-1 transcription and cysl-1 and hif-1 are necessary for survival on high cysteine

Mammalian CDO1 levels and activity are highly induced by dietary cysteine (14–16). To determine if this homeostatic response is conserved in C. elegans, we exposed transgenic C. elegans carrying the Pcdo-1::GFP transcriptional reporter to high supplemental cysteine (100μM). Like our egl-9 and rhy-1 loss-of-function mutations, high cysteine promoted cdo-1 transcription (Fig. 3).

High cysteine activates cdo-1 transcription

A) Expression of the Pcdo-1::GFP transgene is displayed for wild-type, young-adult C. elegans exposed to 100μM supplemental cysteine as well as non-supplemented control animals. Scale bar is 250μm. B) Quantification of the data displayed in Fig. 3A. Individual datapoints are shown (circles) as are the mean and standard deviation (red lines). n is 6 individuals per genotype. Data are normalized so that expression of Pcdo-1::GFP in wild-type C. elegans exposed to 0μM supplemental cysteine is equal to 1 arbitrary unit (a.u.). **, p<0.001, Welch’s t test.

We hypothesized that cysteine might activate cdo-1 transcription through the RHY-1/CYSL-1/EGL-9/HIF-1 pathway. In this pathway, cysl-1 and hif-1 act to promote cdo-1 transcription. We tested whether cysl-1 or hif-1 were necessary for the induction of Pcdo-1::GFP by high cysteine. However, this experiment was not possible as we observed 100% lethality in cysl-1(-); Pcdo-1::GFP (n=174 individuals) and hif-1(-); Pcdo-1::GFP (n=139 individuals) animals when exposed to 100μM supplemental cysteine. In contrast, 100% of cysl-1(-); Pcdo-1::GFP (n=143 individuals) and 99% hif-1(-); Pcdo-1::GFP (n=110 individuals) animals were alive under control conditions without supplemental cysteine. Wild-type C. elegans carrying the Pcdo-1::GFP reporter transgene were healthy under control (100% alive, n=134 individuals) and high-cysteine (100% alive, n=140 individuals) conditions. These data demonstrate that cysl-1 and hif-1 are necessary for survival under high cysteine conditions.

Given the established role of CDO-1 in cysteine catabolism, we tested whether cdo-1(-) mutants were also sensitive to high cysteine. 98% of cdo-1(-) mutants were alive after exposure to 100μM supplemental cysteine (n=141 individuals). 100% of cdo-1(-) mutant animals were alive in the control group not exposed to high cysteine (n=142 individuals). Similarly, wild-type C. elegans were healthy under control (100% alive, n=131 individuals) and high-cysteine (100% alive, n=153 individuals) conditions. Thus, cdo-1 is not necessary for survival under high cysteine conditions. This suggests the existence of alternate pathways that promote cysteine homeostasis. Given the requirement of cysl-1 and hif-1 for survival in high cysteine, we propose that HIF-1 activates pathways (in addition to cdo-1) that promote survival under high cysteine conditions.

Activated CDO-1 accumulates and is functional in the hypodermis

We sought to identify the site of action of CDO-1. To observe CDO-1 localization, we used CRISPR/Cas9 to insert the GFP open reading frame into the endogenous cdo-1 locus, fused in place of the native cdo-1 stop codon. This new cdo-1(rae273) allele encodes a C-terminal tagged “CDO-1::GFP” protein from the native cdo-1 genomic locus (Fig. 4A). To determine if CDO-1::GFP was functional, we combined CDO-1::GFP with a loss-of-function mutation in moc-1, a gene that is essential for C. elegans Moco biosynthesis (33). We then observed the growth of moc-1(-) CDO-1::GFP C. elegans on wild-type and Moco-deficient E. coli. moc-1(-) mutant C. elegans expressing CDO-1::GFP from the native cdo-1 locus arrest during larval development when fed Moco-deficient E. coli, but not when fed Moco-producing E.coli (Fig. 4C). This lethality is caused by the CDO-1-mediated production of sulfites which are only toxic when C. elegans is Moco deficient, and demonstrates that the CDO-1::GFP fusion protein is functional (33). These data suggest that the CDO-1::GFP expression we observe is physiologically relevant.

Hypodermal CDO-1 accumulates in the cytoplasm when egl-9 or rhy-1 are inactive and is sufficient to promote sulfur amino acid metabolism

A) Diagram of cdo-1(rae273), a CRISPR/Cas9-generated allele where GFP is inserted into endogenous cdo-1, creating a C-terminal CDO-1::GFP fusion protein expressed from the native cdo-1 locus. B) Differential interference contrast (DIC) and fluorescence imaging are shown for wild-type, egl-9(sa307), and rhy-1(ok1402) C. elegans expressing CDO-1::GFP encoded by cdo-1(rae273). Scale bar is 10μm. An anterior segment of the Hyp7 hypodermal cell is displayed. CDO-1::GFP accumulates in the cytoplasm and is excluded from the nuclei. C) moc-1(ok366), moc-1(ok366) cdo-1(mg622), and moc-1(ok366) cdo-1(rae273) animals were cultured from synchronized L1 larvae for 72 hours on wild-type (black, Moco+) or ΔmoaA mutant (red, Moco-) E. coli. D) moc-1(ok366) cdo-1(mg622) double mutant animals expressing Pcol-10::CDO-1::GFP or Pcol-10::CDO-1[C85Y]::GFP transgenes were cultured for 48 hours on wild-type (black, Moco+) or ΔmoaA mutant (red, Moco-) E. coli. Two independently derived strains were tested for each transgene. For panels C and D, animal lengths were determined for each condition. Box plots display the median, upper, and lower quartiles and whiskers indicate minimum and maximum data points. Sample size (n) is 10 individuals for each experiment. ****, p<0.0001, multiple unpaired t test with Welch correction. ns indicates no significant difference was identified.

When wild-type animals expressing CDO-1::GFP were grown under standard culture conditions, we observed CDO-1::GFP expression in multiple tissues including prominent expression in the hypodermis (Fig. 4B, Fig. S2). We tested if CDO-1::GFP levels would be affected by inactivation of egl-9 or rhy-1. We generated egl-9(-); CDO-1::GFP and rhy-1(-); CDO-1::GFP animals and assayed expression of CDO-1::GFP. Consistent with our studies using Pcdo-1::CDO-1::GFP and Pcdo-1::GFP transgenes, we found that CDO-1::GFP levels, encoded by cdo-1(rae273), were increased by egl-9(-) or rhy-1(-) mutations (Fig. 4B, Fig. S2). Furthermore, high cysteine also promoted accumulation of CDO-1::GFP (Fig. S3). In all scenarios, the hypodermis was the most prominent site of CDO-1::GFP accumulation. Specifically, we found that CDO-1::GFP was expressed in the cytoplasm of Hyp7, the major C. elegans hypodermal cell (Fig. 4B).

Based on the expression pattern of CDO-1::GFP encoded by cdo-1(rae273), we hypothesized that CDO-1 acts in the hypodermis to promote sulfur amino acid metabolism. To test this hypothesis, we engineered a cdo-1 rescue construct in which cdo-1 is expressed exclusively in the hypodermis under the control of the col-10 promoter (Pcol-10::CDO-1::GFP) (52). Multiple independent transgenic C. elegans strains were generated by integrating the Pcol-10::CDO-1::GFP construct into the C. elegans genome (41). We tested the ability of the Pcol-10::CDO-1::GFP transgene to rescue the cdo-1(-) mutant suppression of Moco-deficient larval arrest. We found that multiple independent transgenic strains of cdo-1(-) moc-1(-) double mutant animals carrying the Pcol-10::CDO-1::GFP transgene displayed a larval arrest phenotype when fed a Moco-deficient diet (Fig. 4D). These data demonstrated that hypodermal-specific expression of cdo-1 is sufficient to rescue the cdo-1(-) mutant suppression of Moco-deficient lethality. This rescue was dependent upon the enzymatic activity of CDO-1 as an active site variant of this transgene (Pcol-10::CDO-1[C85Y]::GFP) did not rescue the suppressed larval arrest of cdo-1(-) moc-1(-) double mutant animals fed Moco-deficient diets (Fig. 4D). Taken together, our analyses of CDO-1::GFP expression demonstrate that CDO-1 is expressed, and that expression is regulated, in multiple tissues, principal among them being the hypodermis and Hyp7 cell. Our tissue-specific rescue data demonstrate that hypodermal expression of cdo-1 is sufficient to promote cysteine catabolism and suggest that the hypodermis is a critical tissue for sulfur amino acid metabolism. However, we cannot exclude the possibility that CDO-1 also acts in other cells and tissues.

HIF-1 promotes CDO-1 activity downstream of the H2S-sensing pathway

We sought to determine the physiological impact of cdo-1 transcriptional activation by HIF-1. We reasoned mutations that activate HIF-1 and increase cdo-1 transcription may cause increased CDO-1 activity. CDO-1 sits at a critical metabolic node in the degradation of the sulfur amino acids cysteine and methionine (Fig 1A). A key byproduct of sulfur amino acid metabolism and CDO-1 is sulfite, a reactive toxin that is detoxified by the Moco-requiring sulfite oxidase (SUOX-1). Null mutations in suox-1 cause larval lethality. However, animals carrying the suox-1(gk738847) hypomorphic allele are healthy under standard culture conditions (33, 34). suox-1(gk738847) mutant animals display only 4% SUOX-1 activity compared to wild type, and are exquisitely sensitive to sulfite stress (53). Thus, the suox-1(gk738847) mutation creates a sensitized genetic background to probe for increases in endogenous sulfite production. To test if increased cdo-1 transcription would impact the growth of suox-1-comprimised animals, we combined the egl-9 null mutation, which promotes HIF-1 activity and cdo-1 transcription, with the suox-1(gk738847) allele. While egl-9(-) and suox-1(gk738847) single mutant animals are healthy under standard culture conditions, the egl-9(-); suox-1(gk738847) double mutant animals are extremely sick and slow growing, establishing a synthetic genetic interaction between these loci (Table 1). To determine the role of sulfur amino acid metabolism in the egl-9(-); suox-1(gk738847) synthetic sick/slow growth phenotype, we engineered egl-9(-); cdo-1(-) suox-1(gk738847) and cth-2(-); egl-9(-); suox-1(gk738847) triple mutant animals. The egl-9; suox-1 synthetic sick/slow growth phenotype was suppressed by inactivating mutations in cdo-1 or cth-2 which block the endogenous production of sulfite (Table 1). These data demonstrate that the deleterious activity of the egl-9(-) mutation in a suox-1(gk738847) background requires functional sulfur amino acid metabolism.

To determine the role of the H2S-sensing pathway in the synthetic sick/slow growth phenotype displayed by egl-9(-); suox-1(gk738847) double mutant animals, we introduced null alleles of cysl-1 or hif-1 into the egl-9(-); suox-1(gk738847) double mutant. The egl-9; suox-1 synthetic sick/slow growth phenotype was dependent upon hif-1 but not cysl-1 (Table 1). These results are consistent with our proposed genetic pathway and support the model that transcriptional activation of cdo-1 by HIF-1 causes increased CDO-1 activity and increased flux of sulfur amino acids through their catabolic pathway.

Loss of rhy-1 also strongly activates cdo-1 transcription. We hypothesized that rhy-1(-); suox-1(gk738847) double mutant animals would display a synthetic sick/slow growth phenotype like egl-9(-); suox-1(gk738847) double mutant animals. However, rhy-1(-); suox-1(gk738847) double mutant animals were just as healthy as either rhy-1(-) or suox-1(gk738847) single mutant C. elegans (Table 1). These data suggest that increasing cdo-1 transcription alone is not sufficient to promote sulfite production via CDO-1. In addition to the role played by rhy-1 in the regulation of HIF-1 activity, rhy-1 itself is a transcriptional target of HIF-1. Loss of egl-9 activity induces rhy-1 mRNA ∼50-fold in a hif-1-dependent manner (49). Given this potent transcriptional activation, we wondered if rhy-1 might play an additional role downstream of HIF-1 in the regulation of sulfur amino acid metabolism and sulfite production. To test this hypothesis, we engineered rhy-1(-); egl-9(-); suox-1(gk738847) triple mutant animals. To our surprise, the rhy-1(-); egl-9(-); suox-1(gk738847) triple mutant animals were healthy, demonstrating that rhy-1 was necessary for the deleterious activity of the egl-9(-) mutation in a suox-1(gk738847) background. These genetic data suggest a dual role for rhy-1 in the control of sulfur amino acid metabolism; first as a component of a regulatory cascade that controls the activity of HIF-1 and second as a functional downstream effector of HIF-1 that is required for sulfur amino acid metabolism.

This is not the first description of a rhy-1 role downstream of hif-1. Overexpression of a rhy-1-encoding transgene suppresses the lethality of a hif-1(-) mutant during HSS stress (54). These data establish RHY-1 as both a regulator and effector of HIF-1. How RHY-1, a predicted membrane-bound O-acyltransferase, molecularly executes these dual roles remains to be explored.

EGL-9 prolyl hydroxylase activity and VHL-1 are largely dispensable in the regulation of CDO-1

EGL-9 inhibits HIF-1 through its prolyl hydroxylase domain (PHD) (Fig. 5A) (39). To evaluate the impact of the EGL-9 PHD on the regulation of cdo-1, we generated a PHD-inactive egl-9 mutation using CRISPR/Cas9. We engineered an egl-9 mutation that substitutes an alanine in place of histidine 487 (H487A) (Fig. 1C). Histidine 487 of EGL-9 is highly conserved and catalytically essential in the PHD active site (Fig. 5B) (55, 56). To evaluate the impact of an inactive EGL-9 PHD on the transcription of cdo-1, we engineered a C. elegans strain carrying the egl-9(H487A) mutation with the Pcdo-1::GFP transcriptional reporter. egl-9(H487A) caused a modest increase in Pcdo-1::GFP accumulation in the C. elegans intestine, suggesting that the EGL-9 PHD is necessary to repress cdo-1 transcription in the intestine (Fig. 5C,D). However, the activation of Pcdo-1::GFP by the egl-9(H487A) mutation was markedly less when compared to Pcdo-1::GFP activation caused by an egl-9(-) null mutation (Fig. 5C,D). These data suggest that EGL-9 has a PHD-independent activity that is responsible for repressing cdo-1 transcription.

egl-9 inhibits cdo-1 transcription in a prolyl-hydroxylase and VHL-1-independent manner

A) The pathway of HIF-1 processing during normoxia is displayed. EGL-9 uses O2 as a substrate to hydroxylate (-OH) HIF-1 on specific proline residues. Prolyl hydroxylated HIF-1 is bound by VHL-1 which facilitates HIF-1 polyubiquitination and targets HIF-1 for degradation by the proteasome. B) Amino acid alignment of the EGL-9 prolyl hydroxylase domain from C. elegans, D. melanogaster, M. musculus, and H. sapiens. “*” indicate perfect amino acid conservation while “.” indicates weak similarity. Highlighted (red) is the catalytically essential histidine 487 residue in C. elegans. Alignment was performed using Clustal Omega (EMBL-EBI). C) Expression of Pcdo-1::GFP transgene is displayed for wild-type, egl-9(sa307, -), vhl-1(ok161), and egl-9(rae276, H487A) C. elegans animals at the L4 stage of development. Scale bar is 250μm. D) Quantification of the data displayed in Fig. 5C. Individual datapoints are shown (circles) as are the mean and standard deviation (red lines). n is 6 individuals per genotype. Data are normalized so that wild-type expression of Pcdo-1::GFP is 1 arbitrary unit (a.u.). ***, p<0.001, ****, p<0.0001, ordinary one-way ANOVA with Tukey’s multiple comparisons test. ns indicates no significant difference was identified.

EGL-9 PHD hydroxylates specific proline residues on HIF-1. These hydroxylated proline residues are recognized by the von Hippel-Lindau E3 ubiquitin ligase (VHL-1). VHL-1-mediated ubiquitination promotes degradation of HIF-1 by the proteasome (Fig. 5A) (40). Thus, the EGL-9 PHD and VHL-1 act in a pathway to regulate HIF-1. To determine the role of VHL-1 in regulating cdo-1, we engineered a C. elegans strain carrying the vhl-1(-) null mutation with our Pcdo-1::GFP reporter. vhl-1 inactivation also caused a modest increase in Pcdo-1::GFP expression in the C. elegans intestine, suggesting that vhl-1 is necessary to repress cdo-1 transcription (Fig. 5C,D). However, activation of Pcdo-1::GFP by the vhl-1(-) mutation was less than activation caused by an egl-9(-) null mutation (Fig. 5C,D). These data suggest that EGL-9 has a VHL-1-independent activity that is responsible for repressing cdo-1 transcription.

To evaluate the impact of inactivating the EGL-9 PHD or VHL-1 on cysteine metabolism, we again employed the suox-1(gk738847) hypomorphic mutation that sensitizes animals to increases in sulfite. We engineered egl-9(H487A); suox-1(gk738847) and vhl-1(-) suox-1(gk738847) double mutant animals and evaluated the health of those strains. In contrast to egl-9(-); suox-1(gk738847) double mutant animals which are extremely sick/slow growing, egl-9(H487A); suox-1(gk738847) and vhl-1(-) suox-1(gk738847) double mutant animals are healthy (Table 1). These genetic data suggest that neither the EGL-9 PHD nor VHL-1 are necessary to repress cysteine catabolism/sulfite production. However, it is plausible that the egl-9(H487) or vhl-1(-) mutations modestly activate cysteine metabolism, likely proportional to their activation of the Pcdo-1::GFP transgene (Fig. 5C,D), and that this activation is not sufficient to produce enough sulfites to negatively impact the growth of suox-1(gk738847) mutant animals.

Discussion

CDO-1 is a physiologically relevant effector of HIF-1

The hypoxia-inducible factor HIF-1 is a master regulator of the cellular response to hypoxia. It activates the transcription of many genes and pathways that are critical to maintain metabolic homeostasis in the face of low oxygen. For example, mammalian HIF1A induces the hematopoietic growth hormone erythropoietin, glucose transport and glycolysis, as well as lactate dehydrogenase (57–60). The nematode C. elegans encounters a range of oxygen tensions in its natural habitat of rotting material: as microbial abundance increases, oxygen levels decrease from atmospheric levels (∼21% O2). In fact, C. elegans prefers 5-12% O2, perhaps because hypoxia predicts abundant bacterial food sources (61). Elements of the HIF-1 pathway have emerged from C. elegans genetics as have critical targets of HIF-1 (39). For instance, C. elegans HIF-1 promotes H2S homeostasis by inducing transcription of the mitochondrial sulfide quinone oxidoreductase (sqrd-1), detoxifying H2S (37).

We sought to define genes that regulate the levels and activity of cysteine dioxygenase (CDO-1), a critical regulator of cysteine homeostasis (14–16). Taking an unbiased genetic approach in C. elegans, we found that cdo-1 was highly regulated by HIF-1 downstream of a signaling pathway that includes rhy-1, cysl-1, and egl-9. We demonstrated that HIF-1 promotes cdo-1 transcription, accumulation of CDO-1 protein, and increased CDO-1 activity. Loss of rhy-1 or egl-9 activate hif-1 to in turn strongly induce the cdo-1 promoter. The pathway for activation of cdo-1 also requires cysl-1, which functions downstream of rhy-1 and upstream of egl-9. Based on ChIP-Seq studies, HIF-1 directly binds to the cdo-1 promoter (50).

We also show that the transcriptional activation of cdo-1 by HIF-1 promotes CDO-1 enzymatic activity. Genetic activation of cdo-1 by loss of egl-9 causes severe sickness in a mutant with reduced sulfite oxidase activity, an activity required to cope with the toxic products of CDO-1. This synthetic sickness is dependent upon an intact HIF-1 signaling pathway and a functioning sulfur amino acid metabolism pathway, as the dramatic sickness of an egl-9; suox-1 double mutant is suppressed by loss of hif-1, cth-2, or cdo-1. These striking genetic results validate the intersection of HIF-1 and CDO-1 as converging biological regulatory pathways and encourage further exploration of this regulatory node. The potential physiological relevance of the connection between HIF-1 signaling and cysteine metabolism is discussed below.

Evidence for distinct pathways of HIF-1 activation by hypoxia and cysteine/H2S

The hypoxia-signaling pathway is defined by EGL-9-dependent prolyl hydroxylation of HIF-1. Hydroxylated HIF-1 is then targeted for degradation via VHL-1-mediated ubiquitination (39, 40, 51). However, multiple lines of evidence, reinforced by our work, demonstrate VHL-1- and prolyl hydroxylase-independent activity of EGL-9. egl-9(-) null mutant C. elegans accumulate HIF-1 protein and display increased transcription of many genes, including nhr-57, an established target of HIF-1 (36). In rescue experiments of an egl-9(-) null mutant, Shao et al. (2009) demonstrate that a wild-type egl-9 transgene restores normal HIF-1 protein levels and HIF-1 transcription. However, rescue experiments with a PHD-inactive egl-9(H487A) transgene do not correct the accumulation of HIF-1 protein and only partially reduce the HIF-1 transcriptional output (56). Through our studies of cdo-1 transcription, we demonstrate that an egl-9(H487A) mutant incompletely activates HIF-1 transcription when compared to an egl-9(-) null mutation (Fig. 5C,D). Taken together, we conclude EGL-9 has activity independent of its PHD domain, mirroring and supporting previous work (56).

In studies of the HIF-1-dependent Pnhr-57::GFP transcriptional reporter, Shen et al. (2006) observed that egl-9(-) null mutants promote HIF-1 transcription more than a vhl-1(-) mutant (36). We observe this same distinction between egl-9 and vhl-1 mutations with our Pcdo-1::GFP reporter (Fig. 5C,D). This difference in transcriptional activity is not explained by HIF-1 protein levels as HIF-1 protein accumulated equally in egl-9(-) and vhl-1(-) null mutants (36). This observation suggests that EGL-9 represses both HIF-1 levels and activity. This study also notes that vhl-1 represses HIF-1 transcription in the intestine while egl-9 acts in a wider array of tissues including the intestine and the hypodermis (36). Budde and Roth (2010) also demonstrate that loss of vhl-1 promotes HIF-1 transcription in the C. elegans intestine while total loss of egl-9 promotes HIF-1 transcription in multiple tissues including the intestine and the hypodermis (62). Our data expand upon these observations by demonstrating that loss of the EGL-9 PHD promotes cdo-1 transcription in the intestine alone, mirroring the loss of vhl-1 (Fig. 5C). Budde and Roth (2010) additionally demonstrate that physiologically relevant stimuli also elicit a tissue-specific transcriptional response: hypoxia promotes HIF-1 transcription in the intestine while H2S promotes HIF-1 transcription in the hypodermis. Importantly, H2S promotes hypodermal HIF-1 transcription in a vhl-1(-) mutant, demonstrating a VHL-1-independent pathway for H2S activation of HIF-1 (62). We demonstrate that high cysteine promotes cdo-1 transcription in the hypodermis. Taken together with our data, these studies suggest 2 distinct pathways for activating HIF-1 transcription: i) a hypoxia-sensing pathway that is dependent upon vhl-1 and the EGL-9 PHD and promotes HIF-1 activity in the intestine and ii) an H2S-sensing pathway that is independent of vhl-1 and the EGL-9 PHD and promotes HIF-1 activity in the hypodermis.

The genetic details of the H2S-sensing pathway were solidified through studies of rhy-1 in C. elegans. Ma et al. (2012) demonstrate that hif-1 repression via rhy-1 requires the activity of cysl-1, a gene encoding a cysteine synthase-like protein (38). They further demonstrate that high H2S promotes a physical interaction between CYSL-1 and EGL-9, resulting in the inactivation of EGL-9 and increased HIF-1 activity. Together, these studies suggest distinct genetic regulators of EGL-9/HIF-1 signaling: rhy-1 and cysl-1 govern the H2S-sensing pathway while vhl-1 mediates the hypoxia-sensing pathway. These pathways are distinct in their requirement for the EGL-9 PHD.

A negative feedback loop senses high cysteine/H2S, promotes CDO-1 activity, and maintains cysteine homeostasis

Why would the RHY-1/CYSL-1/EGL-9/HIF-1 H2S-sensing pathway control the levels and activity of cysteine dioxygenase? We speculate this intersection facilitates a homeostatic pathway allowing C. elegans to sense and respond to cysteine level. We propose that H2S acts as a gaseous signaling molecule to promote cysteine catabolism. H2S activates HIF-1 in the hypodermis by promoting the CYSL-1-mediated inactivation of EGL-9 (38). We show that high cysteine similarly induces cdo-1 transcription in the hypodermis. Our genetic data demonstrate that cdo-1 is induced by the same genetic pathway that senses H2S in C. elegans and CDO-1 acts in the hypodermis, the major site of H2S-induced transcription. Furthermore, H2S induces endogenous cdo-1 transcription >3-fold but cdo-1 mRNA levels do not change when C. elegans are exposed to hypoxia (63, 64). Thus, it is likely that H2S promotes cdo-1 transcription through RHY-1, CYSL-1, EGL-9, and HIF-1. H2S is a reasonable small molecule signal to alert cells to high cysteine stress. Excess cysteine results in the production of H2S mediated by multiple enzymes including cystathionase (CTH), cystathionine ý-synthase (CBS), and 3-mercaptopyruvate sulfurtransferase (MST) (2, 65). We speculate that excess cysteine in C. elegans promotes the enzymatic production of H2S which would activate HIF-1 via the RHY-1/CYSL-1/EGL-9 signaling pathway. In this homeostatic model, H2S-activated HIF-1 would then induce cdo-1 transcription, promoting CDO-1 activity and the catabolism of the high-cysteine trigger (Fig. 6). Supporting this model, cysl-1(-) and hif-1(-) mutant C. elegans cannot survive in a high cysteine environment, demonstrating their central role in promoting cysteine homeostasis.

Model for the regulation of cysteine metabolism by HIF-1

A) Proposed genetic pathway for the regulation of cdo-1. rhy-1, cysl-1, and egl-9 act in a negative-regulatory cascade to control activity of the HIF-1 transcription factor, which activates transcription of cdo-1. B) Under basal conditions (inactivated HIF-1), EGL-9 negatively regulates HIF-1 through 2 distinct pathways; one pathway is dependent upon O2, prolyl hydroxylation (PHD), and VHL-1, while the second acts independently of these canonical factors. Under these conditions cdo-1 transcription is kept at basal levels and cysteine catabolism is not induced. C) During conditions where HIF-1 is activated (high H2S), CYSL-1 directly binds and inhibits EGL-9, preventing HIF-1 inactivation. Active HIF-1 binds the cdo-1 promoter, driving transcription and promoting CDO-1 protein accumulation. High CDO-1 promotes the catabolism of cysteine. HIF-1-induced cysteine catabolism requires the activity of rhy-1.

Members of the H2S-sensing pathway have also been implicated in the C. elegans response to infection (44, 66). Many secreted proteins that mediate intercellular signaling or innate immune responses to pathogens are cysteine-rich and form disulfide bonds in the endoplasmic reticulum before protein secretion (67, 68). Levels of free cysteine may fall during the massive inductions of cysteine-rich secreted proteins in development or immune defense (69). The oxidation of so many cysteines to disulfides in the endoplasmic reticulum might locally lower oxygen levels preventing EGLN1 hydroxylation of HIF1A. Thus, the intersection between oxygen sensing and cysteine homeostasis we have uncovered may contribute to a regulatory axis in cell-cell signaling during development and in immune function.

Methods

General methods and strains

C. elegans were cultured using established protocols (43). Briefly, animals were cultured at 20°C on nematode growth media (NGM) seeded with wild-type E. coli (OP50). The wild-type strain of C. elegans was Bristol N2. Additional E. coli strains used in this work were B2W5113 (Wild type, Moco+) and JW0764-2 (τιmoaA753::kan, Moco-) (70).

C. elegans mutant and transgenic strains used in this work are listed here. When previously published, sources of strains are referenced. Unless a reference is provided, all strains were generated in this study.

Non-transgenic C. elegans

JT307, egl-9(sa307) (44)

GR2254, moc-1(ok366) X (33)

GR2260, cdo-1(mg622) X (33)

GR2261, cdo-1(mg622) moc-1(ok366) X (33)

GR2269, suox-1(gk738847) X (33)

CB5602, vhl-1(ok161) X (39)

USD414, rhy-1(ok1402) II; suox-1(gk738847) X

USD421, egl-9(sa307) V; suox-1(gk738847) X

USD422, vhl-1(ok161) suox-1(gk738847) X

USD430, egl-9(sa307) V; cdo-1(mg622) suox-1(gk738847) X

USD431, hif-1(ia4) egl-9(sa307) V; suox-1(gk738847) X

USD432, rhy-1(ok1402) II; egl-9(sa307) V; suox-1(gk738847) X

USD433, cth-2(mg599) II; egl-9(sa307) V; suox-1(gk738847) X

USD434, egl-9(sa307) X; cysl-1(ok762) suox-1(gk738847) X

USD512, rhy-1(ok1402) II Outcrossed 4x for this work

USD706, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X

USD920, cdo-1(rae273) moc-1(ok366) X

USD921, egl-9(sa307) V; cdo-1(rae273) X

USD922, rhy-1(ok1402) II; cdo-1(rae273) X

USD937, egl-9(rae276) V; suox-1(gk738847) X

MiniMos transgenic lines

USD531, unc-119(ed3) III; raeTi1 [pcdo-1::CDO-1::GFP unc-119(+)]

USD719, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366); raeTi14 [pcdo-1::CDO-1(C85Y)::GFP unc-119(+)]

USD720, unc-119(ed3) III; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD730, rhy-1(ok1402) II; unc-119(ed3) III; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD733, unc-119(ed3) III; egl-9(sa307) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD739, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi1 [pcdo-1::CDO-1::GFP unc-119(+)]

USD766, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi32 [Pcol-10::CDO-1::GFP unc-119(+)]

USD767, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi33 [Pcol-10::CDO-1::GFP unc-119(+)]

USD776, rhy-1(ok1402) II; unc-119(ed3) III; hif-1(ia4) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD777, unc-119(ed3) III; egl-9(sa307) hif-1(ia4) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD780, rhy-1(ok1402) II; unc-119(ed3) III; cysl-1(ok762) X; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD787, unc-119(ed3) III; egl-9(sa307) V; cysl-1(ok762) X; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD808, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi40 [Pcol-10::CDO-1[C85Y]::GFP unc-119(+)]

USD810, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi41 [Pcol-10::CDO-1[C85Y]::GFP unc-119(+)]

USD940, unc-119(ed3) III; vhl-1(ok161) X; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD1160, unc-119(ed3) III; cysl-1(ok762) X; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD1161, unc-119(ed3) III; hif-1(ia4) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

EMS-derived strains

USD659, unc-119(ed3) III; egl-9(rae213) V; raeTi1 [pcdo-1::CDO-1::GFP unc-119(+)]

USD674, unc-119(ed3) III; egl-9(rae227) V; raeTi1 [pcdo-1::CDO-1::GFP unc-119(+)]

USD655, rhy-1(rae209) II; unc-119(ed3) III; raeTi1 [pcdo-1::CDO-1::GFP unc-119(+)]

USD656, rhy-1(rae210) II; unc-119(ed3) III; raeTi1 [pcdo-1::CDO-1::GFP unc-119(+)]

USD657, rhy-1(rae211) II; unc-119(ed3) III; raeTi1 [pcdo-1::CDO-1::GFP unc-119(+)]

USD658, rhy-1(rae212) II; unc-119(ed3) III; raeTi1 [pcdo-1::CDO-1::GFP unc-119(+)]

CRISPR/Cas9-derived strains

USD914, cdo-1(rae273) X, CDO-1::GFP

USD926, egl-9(rae276) V, EGL-9[H487A]

USD928, unc-119(ed3) III; egl-9(rae278) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

MiniMos transgenesis

Cloning of original plasmid constructs was performed using isothermal/Gibson assembly (71). All MiniMos constructs were assembled in pNL43, which is derived from pCFJ909, a gift from Erik Jorgensen (Addgene plasmid #44480) (72). Details about plasmid construction are described below. MiniMos transgenic animals were generated using established protocols that rescue the unc-119(ed3) Unc phenotype (41).

To generate a construct that expressed CDO-1 under the control of its native promoter, we cloned the wild-type cdo-1 genomic locus from 1,335 base pairs upstream of the cdo-1 ATG start codon to (and including) codon 190 encoding the final CDO-1 amino acid prior to the TAA stop codon. This wild-type genomic sequence was fused in frame with a C-terminal GFP and tbb-2 3’UTR (376 bp downstream of the tbb-2 stop codon) in the pNL43 plasmid backbone. This plasmid is called pKW24 (Pcdo-1::CDO-1::GFP).

To generate pKW44, a construct encoding the active site mutant transgene Pcdo-1::CDO-1(C85Y)::GFP, we performed Q5 site-directed mutagenesis on pKW24, following manufacturer’s instructions (New England Biolabs). In pKW44, codon 85 was mutated from TGC (cysteine) to TAC (tyrosine).

To generate pKW45 (Pcdo-1::GFP), the 1,335 base pair cdo-1 promoter was amplified and fused directly to the GFP coding sequence. Both fragments were derived from pKW24, excluding the cdo-1 coding sequence and introns.

pKW49 is a construct driving cdo-1 expression from the hypodermal-specific col-10 promoter (Pcol-10::CDO-1::GFP) (52). The col-10 promoter (1,126 base pairs upstream of the col-10 start codon) was amplified and fused upstream of the cdo-1 ATG start codon in pKW24, replacing the native cdo-1 promoter. pKW53 [Pcol-10::CDO-1(C85Y)::GFP] was engineered using the same Q5-site-directed mutagenesis strategy as was described for pKW44. However, this mutagenesis used pKW49 as the template plasmid.

Chemical mutagenesis and whole genome sequencing

To define C. elegans gene activities that were necessary for the control of cdo-1 levels, we carried out a chemical mutagenesis screen for animals that accumulate CDO-1 protein. To visualize CDO-1 levels, we engineered USD531, a transgenic C. elegans strain carrying the raeTi1 [pKW24, Pcdo-1::CDO-1::GFP] transgene. USD531 transgenic C. elegans were mutagenized with ethyl methanesulfonate (EMS) using established protocols (43). F2 generation animals were manually screened, and mutant isolates were collected that displayed high accumulation of Pcdo-1::CDO-1::GFP. We demanded that mutant strains of interest were viable and fertile.

We followed established protocols to identify EMS-induced mutations in our strains of interest (73). Briefly, whole genomic DNA was prepared from C. elegans using the Gentra Puregene Tissue Kit (Qiagen) and genomic DNA libraries were prepared using the NEBNext genomic DNA library construction kit (New England Biolabs). DNA libraries were sequenced on an Illumina Hi-Seq and deep sequencing reads were analyzed using standard methods on Galaxy, a web-based platform for computational analyses (74). Briefly, sequencing reads were trimmed and aligned to the WBcel235 C. elegans reference genome (75, 76). Variations from the reference genome and the putative impact of those variations were annotated and extracted for analysis (77–79). Here we report the analysis of 6 new mutant strains (USD655, USD656, USD657, USD658, USD659, and USD674). Among these mutant strains, we found 2 unique mutations in egl-9 (USD659 and USD674) and 4 unique mutations in rhy-1 (USD655, USD656, USD657, USD658). The allele names and molecular identity of these new egl-9 and rhy-1 mutations are specified in Fig. 1C. These genes were prioritized based on the isolation of multiple independent alleles and their established functions in a common pathway, the hypoxia and H2S-sensing pathway (36–38).

Genome engineering by CRISPR/Cas9

We followed standard protocols to perform CRISPR/Cas9 genome engineering of cdo-1 and egl-9 genomic loci in C. elegans (80–83). Essential details of the CRISPR/Cas9-generated reagents in this work are described below.

We used homology-directed repair to generate cdo-1(rae273) [CDO-1::GFP]. The guide RNA was 5’-gactacagaggatctaagaa-3’ (crRNA, IDT). The GFP donor double-stranded DNA (dsDNA) was amplified from pCFJ2249 using primers that contained roughly 40bp of homology to cdo-1 flanking the desired insertion site (84). The primers used to generate the donor dsDNA were: 5’-gtacggcaagaaagttgactacagaggatctaagaataatagtactagcggtggcagtgg-3’ and 5’-agaatcaacacgttattacattgagggatatgttgtttacttgtagagctcgtccattcc-3’. Glycine 189 of CDO-1 was removed by design to eliminate the PAM site and prevent cleavage of the donor dsDNA.

The egl-9(rae276) and egl-9(rae278) [EGL-9(H487A)] alleles were also generated by homology-directed repair using the same combination of guide RNA and single-stranded oligodeoxynucleotide (ssODN) donor. The guide RNA was 5’-tgtgaagcatgtagataatc-3’ (crRNA, IDT) The ssODN donor was 5’-gcttgccatctatcctggaaatggaactcgttatgtgaaggctgtagacaatccagtaaaagatggaagatgtataaccactatttattactg-3’ (Ultramer, IDT). Successful editing resulted in altering the coding sequence of EGL-9 to encode for an alanine rather than the catalytically essential histidine at position 487. We also used synonymous mutations to introduce an AccI restriction site that is helpful for genotyping the rae276 and rae278 mutant alleles.

C. elegans growth assays

To assay developmental rates, C. elegans were synchronized at the first stage of larval development. To synchronize animals, embryos were harvested from gravid adult animals via treatment with a bleach and sodium hydroxide solution. Embryos were then incubated overnight in M9 solution causing them to hatch and arrest development at the L1 stage (85). Synchronized L1 animals were cultured on NGM seeded with B2W5113 (Wild type, Moco+) or JW0764-2 (τ..moaA753::kan, Moco-) E. coli. Animals were cultured for 48 or 72 hours (specified in the appropriate figure legends), and live animals were imaged as described below. Animal length was measured from tip of head to the end of the tail.

To determine qualitative ‘health’ of various C. elegans strains, we assayed the ability of these strains to consume all E. coli food provided on an NGM petri dish. For this experiment, dietary E. coli was produced via overnight culture in liquid LB in a 37°C shaking incubator. 200μl of this E. coli was seeded onto NGM petri dishes and allowed to dry, producing nearly identical lawns and growth environments. Then, 5 L4 animals of a strain of interest were introduced onto these NGM petri dishes seeded with OP50. For each experiment, petri dishes were monitored daily and scored when all E. coli was consumed by the population of animals. This assay is beneficial because it integrates many life-history measures (i.e. developmental rate, brood size, embryonic viability, etc.) into a single simple assay that can be scaled and applied to many C. elegans strains in parallel.

Cysteine exposure

To determine the impact of supplemental cysteine on expression of cdo-1 and animal viability, we exposed various C. elegans strains to 0 or 100μM supplemental cysteine. Wild-type and mutant animals carrying the Pcdo-1::GFP or CDO-1::GFP transgenes were synchronously grown on NGM media supplemented with E. coli OP50 at 20°C until reaching the L4 stage of development. Live L4 animals were then collected from the petri dishes and cultured in liquid M9 media containing 4X concentrated E. coli OP50 with or without 100μM supplemental cysteine. These liquid cultures were gently rocked at 20°C overnight. Post exposure, GFP imaging was performed as described in the Microscopy section of the materials and methods. A subset of the assayed animals were also scored for viability after being seeded onto NGM petri dishes. Animals were determined to be alive if they responded to mechanical stimulus.

Microscopy

Low magnification bright field and fluorescence images (imaging GFP simultaneously in multiple animals) were collected using a Zeiss AxioZoom V16 microscope equipped with a Hamamatsu Orca flash 4.0 digital camera using Zen software (Zeiss). For experiments with supplemental cysteine, low magnification bright field and fluorescence images were collected using a Nikon SMZ25 microscope equipped with a Hamamatsu Orca flash 4.0 digital camera using NIS-Elements software (Nikon). High magnification differential interference contrast (DIC) and GFP fluorescence images (imaging CDO-1::GFP encoded by cdo-1(rae273)) were collected using Zeiss AxioImager Z1 microscope equipped with a Zeiss AxioCam HRc digital camera using Zen software (Zeiss). All images were processed and analyzed using ImageJ software (NIH). All imaging was performed on live animals paralyzed using 20mM sodium azide. For all fluorescence images shown within the same figure panel, images were collected using the same exposure time and processed identically. To quantify GFP expression, the average pixel intensity was determined within a set transverse section immediately anterior to the developing vulva. Background pixel intensity was determined in a set region of interest distinct from the C. elegans samples and was subtracted from the sample measurements.

eLife Assessment

This valuable study presents findings on how the hypoxia response pathway senses and responds to changes in the homeostasis of the amino acid cysteine and other sulfur-containing molecules, with compelling and rigorous genetic analyses. The works adds to a growing body of literature showing that prolyl hydroxylation is not the only mechanism by which the hypoxia response pathway can act. Although the paper does not reveal new biochemical insight into the mechanism, it opens up new areas of investigation that will be of interest to cell biologists and biomedical researchers studying the many pathologies involving hypoxia and/or cysteine metabolism.

Reviewer #1 (Public Review):

Warnhoff et al present a genetic investigation of the response of C. elegans to high dietary cysteine. Using a Pcdo-1::CDO-1::GFP reporter (for a cysteine dioxygenase gene) and unbiased mutagenesis, they identify multiple alleles, including nonsense alleles, in egl-9 and rhy-1, which they validate with reference alleles. Further mutational analysis shows that hif-1 and cysl-1, components of the same established genetic regulatory pathway, also act in cdo-1 regulation. High dietary levels of cysteine activate cdo-1 expression, but loss of cdo-1 does not cause sensitivity to excess dietary cysteine, whereas cysl-1 and hif-1 are completely inviable in these conditions. Using sulfite oxidase suox-1 mutant and double and triple mutant analysis the authors show that the defects caused by suox-1 deletion (which causes sulfite accumulation) are exacerbated by loss of egl-9, which is alleviated by concomitant loss of enzymes cdo-1 / cth-2 or regulators rhy-1 / hif-1, demonstrating that the key issue is cysteine derived sulfites. Further genetic analysis shows that although egl-9 is required for cdo-1 induction, this is only partially dependent on its hydroxylase activity and the egl-9 partner vhl-1 is also only partially involved.

The significance of the findings is that they describe a regulatory pathway by which organisms might respond to high levels of cysteine in vivo.

Strengths

- The genetic analysis is generally well done and convincing, with multiple alleles identified for each gene, several reporters used for cdo-2, etc.

- Genetic analysis using site-directed mutagenesis of egl-9 and cdo-1 with point mtuations is especially nice.

- The data are analyzed and represented properly, and microscopy data have been quantified.

- The paper is also written quite clearly and the figures are easy to understand.

Weaknesses

- The relevance is somewhat unclear. High cysteine levels can be achieved in the laboratory, but, is this relevant in the life of C. elegans? Or is there physiological relevance in humans, e.g. a disease? The authors state "cells and animals fed excess cysteine and methionine", but is this more than a laboratory excess condition? SUOX nonfunctional conditions in humans don't appear to tie into this, since, in that context, the goal is to inactivate CDO or CTH to prevent sulfite production. The authors also mention cancer, but the link to cysteine levels is unclear. In that sense, then, the conditions studied here may not carry much physiological relevance.

- The pathway is described as important for cysteine detoxification, which is described to act via H2S (Figure 6). Much of that pathway has already been previously established by the Roth, Miller, and Horvitz labs as critical for the H2S response. While the present manuscript adds some additional insight such as the additional role of RHY-1 downstream on HIF-1 in promoting toxicity, this study therefore mainly confirms the importance of a previously described signalling pathway, essentially adding a new downstream target rhy-1 -> cysl-1 -> egl-9 -> hif-1 -> sqrd-1/cdo-1. The impact of this finding is reduced by the fact that cdo-1 itself isn't actually required for survival in high cysteine, suggesting it is merely a maker of the activity of this previously described pathway.

Reviewer #2 (Public Review):

The authors investigate the transcriptional regulation of cysteine dioxygenase (CDO-1) in C. elegans and its role in maintaining cysteine homeostasis. They show that high cysteine levels activate cdo-1 transcription through the hypoxia-inducible transcription factor HIF-1. Using transcriptional and translational reporters for CDO-1, the authors propose a negative feedback pathway involving RHY-1, CYSL-1, EGL-9, and HIF-1 in regulating cysteine homeostasis.

Genetics is a notable strength of this study. The forward genetic screen, gene interaction, and epistasis analyses are beautifully designed and rigorously conducted, yielding solid and unambiguous conclusions on the genetic pathway regulating CDO-1. The writing is clear and accessible, contributing to the overall high quality of the manuscript.

Addressing the specifics of cysteine supplementation and interpretation regarding the cysteine homeostasis pathway would further clarify the paper and strengthen the study's conclusions.

First, the authors show that the supplementation of exogenous cysteine activates cdo-1p::GFP. Rather than showing data for one dose, the author may consider presenting dose-dependency results and whether cysteine activation of cdo-1 also requires HIF-1 or CYSL-1, which would be important data given the focus and major novelty of the paper in cysteine homeostasis, not the cdo-1 regulatory gene pathway. While the genetic manipulation of cdo-1 regulators yields much more striking results, the effect size of exogenous cysteine is rather small. Does this reflect a lack of extensive condition optimization or robust buffering of exogenous/dietary cysteine? Would genetic manipulation to alter intracellular cysteine or its precursors yield similar or stronger effect sizes?

Second, there remain several major questions regarding the interpretation of the cysteine homeostasis pathway. How much specificity is involved for the RHY-1/CYSL-1/EGL-9/HIF-1 pathway to control cysteine homeostasis? Is the pathway able to sense cysteine directly or indirectly through its metabolites or redox status in general? Given the very low and high physiological concentrations of intracellular cysteine and glutathione (GSH, a major reserve for cysteine), respectively, there is a surprising lack of mention and testing of GSH metabolism. In addition, what are the major similarities and differences of cysteine homeostasis pathways between C. elegans and other systems (HIF dependency, transcription vs post-transcriptional control)? These questions could be better discussed and noted with novel findings of the current study that are likely C. elegans specific or broadly conserved.

Reviewer #3 (Public Review):

There has been a long-standing link between the biology of sulfur-containing molecules (e.g., hydrogen sulfide gas, the amino acid cysteine, and its close relative cystine, et cetera) and the biology of hypoxia, yet we have a poor understanding of how and why these two biological processes and are co-regulated. Here, the authors use C. elegans to explore the relationship between sulfur metabolism and hypoxia, examining the regulation of cysteine dioxygenase (CDO1 in humans, CDO-1 in C. elegans), which is critical to cysteine catabolism, by the hypoxia inducible factor (HIF1 alpha in humans, HIF-1 in C. elegans), which is the key terminal effector of the hypoxia response pathway that maintains oxygen homeostasis. The authors are trying to demonstrate that (1) the hypoxia response pathway is a key regulator of cysteine homeostasis, specifically through the regulation of cysteine dioxygenase, and (2) that the pathway responds to changes in cysteine homeostasis in a mechanistically distinct way from how it responds to hypoxic stress.

Briefly summarized here, the authors initiated this study by generating transgenic animals expressing a CDO-1::GFP protein chimera from the cdo-1 promoter so that they could identify regulators of CDO-1 expression through a forward genetic screen. This screen identified mutants with elevated CDO-1::GFP expression in two genes, egl-9 and rhy-1, whose wild-type products are negative regulators of HIF-1, raising the possibility that cdo-1 is a HIF-1 transcriptional target. Indeed, the authors provide data showing that cdo-1 regulation by EGL-9 and RHY-1 is dependent on HIF-1 and that regulation by RHY-1 is dependent on CYSL-1, as expected from other published findings of this pathway. The authors show that exogenous cysteine activates cdo-1 expression, reflective of what is known to occur in other systems. Moreover, they find that exogenous cysteine is toxic to worms lacking CYSL-1 or HIF-1 activity, but not CDO-1 activity, suggesting that HIF-1 mediates a survival response to toxic levels of cysteine and that this response requires more than just the regulation of CDO-1. The authors validate their expression studies using a GFP knockin at the cdo-1 locus, and they demonstrate that a key site of action for CDO-1 is the hypodermis. They present genetic epistasis analysis supporting a role for RHY-1, both as a regulator of HIF-1 and as a transcriptional target of HIF-1, in offsetting toxicity from aberrant sulfur metabolism. The authors use CRISPR/Cas9 editing to mutate a key amino acid in the prolyl hydroxylase domain of EGL-9, arguing that EGL-9 inhibits CDO-1 expression through a mechanism that is largely independent of the prolyl hydroxylase activity.

Overall, the data seem rigorous, and the conclusions drawn from the data seem appropriate. The experiments test the hypothesis using logical and clever molecular genetic tools and design. The sample size is a bit lower than is typical for C. elegans papers; however, the experiments are clearly not underpowered, so this is not an issue. The paper is likely to drive many in the field (including the authors themselves) into deeper experiments on (1) how the pathway senses hypoxia and sulfur/cysteine/H2S using these distinct mechanisms/modalities, (2) how oxygen and sulfur/cysteine/H2S homeostasis influence one another, and (3) how this single pathway evolved to sense and respond to both of these stress modalities.

Major strengths of the paper include (1) the use of the powerful whole animal C. elegans model to reveal results that have meaning in vivo, (2) the careful demonstration through mutant rescue experiments that key transgenes have functional activity, (3) the use of CRISPR/Cas9 editing to mutate a critical residue in the catalytic domain of the EGL-9 prolyl hydroxylase, (4) transgenic rescue experiments that show that CDO-1 operates in the hypodermis with regard to the larval arrest phenotype, and (5) the thorough epistatic analysis of different pathway mutants.

Major weaknesses of the paper include (1) the over-reliance on genetic approaches, (2) the lack of novelty regarding prolyl hydroxylase-independent activities of EGL-9, and (3) the lack of biochemical approaches to probe the underlying mechanism of the prolyl hydroxylase-independent activity of EGL-9.

Major Issues We Feel the Authors Should Address:

1. One particularly glaring concern is that the authors really do not know the extent to which the prolyl hydroxylase activity is (or is not) impacted by the H487A mutation in egl-9(rae276). If there is a fair amount of enzymatic activity left in this mutant, then it complicates interpretation. The paper would be strengthened if the authors could show that the egl-9(rae276) eliminates most if not all prolyl hydroxylase activity. In addition, the authors may want to consider doing RNAi for egl-9 in the egl-9(rae276) mutant as a control, as this would support the claim that whatever non-hydroxylase activity EGL-9 may have is indeed the causative agent for the elevation of CDO-1::GFP. Without such experiments, readers are left with the nagging concern that this allele is simply a hypomorph for the single biochemical activity of EGL-9 (i.e., the prolyl hydroxylase activity) rather than the more interesting, hypothesized scenario that EGL-9 has multiple biochemical activities, only one of which is the prolyl hydroxylase activity.

2. The authors observed that EGL-9 can inhibit HIF-1 and the expression of the HIF-1 target cdo-1 through a combination of activities that are (1) dependent on its prolyl hydroxylase activity (and subsequent VHL-1 activity that acts on the resulting hydroxylated prolines on HIF-1), and (2) independent of that activity. This is not a novel finding, as the authors themselves carefully note in their Discussion section, as this odd phenomenon has been observed for many HIF-1 target genes in multiple publications. While this manuscript adds to the description of this phenomenon, it does not really probe the underlying mechanism or shed light on how EGL-9 has these dual activities. This limits the overall impact and novelty of the paper.

3. Cysteine dioxygenases like CDO-1 operate in an oxygen-dependent manner to generate sulfites from cysteine. CDO-1 activity is dependent upon availability of molecular oxygen; this is an unexpected characteristic of a HIF-1 target, as its very activation is dependent on low molecular oxygen. Authors neither address this in the text nor experimentally, and it seems a glaring omission.

4. The authors determined that the hypodermis is the site of the most prominent CDO-1::GFP expression, relevant to Figure 4. This claim would be strengthened if a negative control tissue, in the animal with the knockin allele, were shown. The hypodermal specific expression is a highlight of this paper, so it would make this article even stronger if they could further substantiate this claim.

Minor issues to note:

Mutants for hif-1 and cysl-1 are sensitive to exogenous cysteine levels, yet loss of CDO-1 expression is not sufficient to explain this phenomenon, suggesting other targets of HIF-1 are involved. Given the findings the authors (and others) have had showing a role for RHY-1 in sulfur amino acid metabolism, shouldn't the authors consider testing rhy-1 mutants for sensitivity to exogenous cysteine?

The cysteine exposure assay was performed by incubating nematodes overnight in liquid M9 media containing OP50 culture. The liquid culture approach adds two complications: (1) the worms are arguably starving or at least undernourished compared to animals grown on NGM plates, and (2) the worms are probably mildly hypoxic in the liquid cultures, which complicates the interpretation.

An easily addressable concern is the wording of one of the main conclusions: that cdo-1 transcription is independent of the canonical prolyl hydroxylase function of EGL-9 and is instead dependent on one of EGL-9's non-canonical, non-characterized functions. There are several points in which the wording suggests that CDO-1 toxicity is independent of EGL-9. In their defense, the authors try to avoid this by saying, "EGL-9 PHD," to indicate that it is the prolyl hydroxylase function of EGL-9 that is not required for CDO-1 toxicity. However, this becomes confusing because much of the field uses PHD and EGL-9/EGLN as interchangeable protein names. The authors need to be clear about when they are describing the prolyl hydroxylase activity of EGL-9 rather than other (hypothesized) activities of EGL-9 that are independent of the prolyl hydroxylase activity.

The authors state in the text, "the egl-9; suox-1 double mutants are extremely sick and slow growing." We appreciate that their "health" assay, based on the exhaustion of food from the plate, is qualitative. We also appreciate that it is a functional measure of many factors that contribute to how fast a population of worms can grow, reproduce, and consume that lawn of food. However, unless they do a lifespan assay and/or measure developmental timing and specifically determine that the double mutant animals themselves are developing and/or growing more slowly, we do not think it is appropriate to use the words "slow growing" to describe the population. As they point out, the rate of consumption of food on the plate in their health assay is determined by a multitude and indeed a confluence of factors; the growth rate is one specific one that is commonly measured and has an established meaning.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Issue 1: The relevance is somewhat unclear. High cysteine levels can be achieved in the laboratory, but, is this relevant in the life of C. elegans? Or is there physiological relevance in humans, e.g. a disease? The authors state "cells and animals fed excess cysteine and methionine", but is this more than a laboratory excess condition? SUOX nonfunctional conditions in humans don't appear to tie into this, since, in that context, the goal is to inactivate CDO or CTH to prevent sulfite production. The authors also mention cancer, but the link to cysteine levels is unclear. In that sense, then, the conditions studied here may not carry much physiological relevance.

Response 1: We set out to answer a fundamental question: what pathways regulate the function of cysteine dioxygenase, a highly conserved enzyme in sulfur amino acid metabolism? In an unbiased genetic screen that sampled millions of EMS generated mutations across all ~20,000 C. elegans genes, we discovered loss of function/null mutations in egl-9 and rhy-1, two negative regulators of the hypoxia inducible transcription factor (hif-1). Genetic ablation of the egl-9 or rhy-1 loci are likely not relevant to the life of a C. elegans animal, i.e. this is not representative of a natural state. Yet, this extreme genetic intervention has taught us a new fundamental truth about the interaction between EGL-9/RHY-1, HIF-1, and the transcriptional activation of cdo1. Similarly, the high cysteine levels used in our assays may or may not be representative of a state in nature, we do not know (nor do we make any claims about the environmental relevance of our choice of cysteine concentrations). It seems very plausible that pathological states exist where cysteine concentrations may rise to comparable levels in our experimental system. More importantly, we have started with excess to physiology to elicit a clear response that we can study in the lab. Similar strategies established the cysteine-induction phenotype of CDO1 in mammalian systems. For instance, in Kwon and Stipanuk 2001, hepatocytes are cultured in media supplemented with 2mmol/L cysteine to promote a ~4-fold increase in CDO1 mRNA.

Issue 2: The pathway is described as important for cysteine detoxification, which is described to act via H2S (Figure 6). Much of that pathway has already been previously established by the Roth, Miller, and Horvitz labs as critical for the H2S response. While the present manuscript adds some additional insight such as the additional role of RHY-1 downstream on HIF-1 in promoting toxicity, this study therefore mainly confirms the importance of a previously described signalling pathway, essentially adding a new downstream target rhy-1 -> cysl-1 -> egl-9 -> hif-1 -> sqrd-1/cdo-1. The impact of this finding is reduced by the fact that cdo-1 itself isn't actually required for survival in high cysteine, suggesting it is merely a maker of the activity of this previously described pathway.

Response 2: We agree that the primary impact of our manuscript is the establishment of a novel intersection between the H2S-sensing pathway (largely worked out by Roth, Miller, and Horvitz) and our gene of interest, cysteine dioxygenase. We believe that the connection between these two pathways is exciting as it suggests a logical homeostatic circuit. High cysteine yields enzymatically produced H2S. This H2S may then act as a signal promoting HIF-1 activity (via RHY-1/CYSL-1/EGL-9). High HIF-1 activity increases cdo-1 transcription and activity promoting the degradation of the high-cysteine trigger. As pointed out by the reviewer, cdo-1(-) loss of function alone does not cause cysteine sensitivity at the concentrations tested. Given that cysl-1(-) and hif-1(-) mutants are exquisitely sensitive to high levels of cysteine, we propose that HIF-1 activates the transcription of additional genes that are required for high cysteine tolerance. However, our genetic data show that cdo-1 is more than simply a marker of HIF-1 transcription. Our genetic data in Table 1 demonstrate that HIF-1 activation (caused by egl-9(-)) is sufficient to cause severe sickness in a suox-1 hypomorphic mutant which cannot detoxify sulfites, a critical product of cysteine catabolism. This severe sickness can be reversed by inactivating hif-1, cth-2, or cdo-1. These data demonstrate a functional intersection between the established H2S-sensing pathway and cysteine catabolism governed by cdo-1.

Reviewer #2 (Public Review):

Issue 3: First, the authors show that the supplementation of exogenous cysteine activates cdo-1p::GFP. Rather than showing data for one dose, the author may consider presenting dose-dependency results and whether cysteine activation of cdo-1 also requires HIF-1 or CYSL-1, which would be important data given the focus and major novelty of the paper in cysteine homeostasis, not the cdo-1 regulatory gene pathway.

Response 3: We agree with the reviewer and have performed the suggested dose-response curve for expression of Pcdo-1::GFP in wild-type C. elegans. We observe substantial activation of the Pcdo-1::GFP transcriptional reporter beginning at 100µM supplemental cysteine (Figure 3C). Higher doses of cysteine do not elicit a substantially stronger induction of the Pcdo-1::GFP reporter. Thus, we find that 100µM supplemental cysteine strikes the right balance between strongly inducing the Pcdo-1::GFP reporter while not inducing any toxicity or lethality in wild-type animals (Figure 3E).

We further agree that testing for induction of the Pcdo-1::GFP reporter in a hif-1(-) or cysl-1(-) mutant background is a critical experiment. However, we have not been able to identify a cysteine concentration that induces Pcdo-1::GFP and is not 100% lethal for hif-1(-) or cysl-1(-) mutant C. elegans. The remarkable sensitivity of hif-1(-) or cysl-1(-) mutant C. elegans to supplemental cysteine demonstrates the critical role of these genes in promoting cysteine homeostasis. But because of this lethality, we could not assay the Pcdo1::GFP reporter in the hif-1(-) or cysl-1(-) mutant animals. But the lethality to excess cysteine demonstrates that this cysteine response is salient. To get at how cysteine might be interacting with the HIF-1-signaling pathway, we performed new additivity experiments by supplementing 100µM cysteine to wild type, egl-9(-), and rhy-1(-) mutant C. elegans expressing the Pcdo-1::GFP reporter. Surprisingly, we found that cysteine had no significant impact on Pcdo-1::GFP expression in an egl-9(-) mutant background but significantly increased the Pcdo-1::GFP expression in a rhy-1(-) background (Figure 3A,B). These data suggest that cysteine acts in a pathway with egl-9 and in parallel to rhy-1. These data have been incorporated into Figure 3A,B and are included in the Results section of the manuscript.

Issue 4: While the genetic manipulation of cdo-1 regulators yields much more striking results, the effect size of exogenous cysteine is rather small. Does this reflect a lack of extensive condition optimization or robust buffering of exogenous/dietary cysteine? Would genetic manipulation to alter intracellular cysteine or its precursors yield similar or stronger effect sizes?

Response 4: We agree that the induction of the Pcdo-1::GFP reporter by supplemental cysteine is not as dramatic as the induction caused by the egl-9 or rhy-1 null alleles. We believe our Response 3 and new Figure 3C demonstrate that this phenomenon is not due to lack of condition optimization, but likely reflects some biology. As pointed out by the reviewer, C. elegans likely buffers exogenous cysteine and this (perhaps) prevents the impressive Pcdo-1::GFP induction observed in the egl-9(-) and rhy-1(-) mutant animals. We have now mentioned this possible interpretation in the Results section. Furthermore, we like the idea of using genetic tricks to promote cysteine accumulation within C. elegans cells and tissues and will consider these approaches in future studies.

Issue 5: Second, there remain several major questions regarding the interpretation of the cysteine homeostasis pathway. How much specificity is involved for the RHY-1/CYSL-1/EGL-9/HIF-1 pathway to control cysteine homeostasis? Is the pathway able to sense cysteine directly or indirectly through its metabolites or redox status in general? Given the very low and high physiological concentrations of intracellular cysteine and glutathione (GSH, a major reserve for cysteine), respectively, there is a surprising lack of mention and testing of GSH metabolism.

Response 5: Future studies are required to determine the specificity of the RHY-1/CYSL-1/EGL-9/HIF-1 pathway for the control of cysteine homeostasis. Our proposed mechanism, that H2S activates the HIF-1 pathway is based largely on the work of the Horvitz lab (Ma et al. 2012). They demonstrate that H2S promotes a direct inhibitory interaction between CYSL-1 and EGL-9, leading to activation of HIF-1. These findings align nicely with our genetic and pharmacological data. However, our work does not provide direct evidence as to the cysteine-derived metabolite that activates HIF-1. We propose H2S as a likely candidate.

We have added a note to the introduction regarding the role of GSH as a reservoir of excess cysteine and agree that future studies might find interesting links between CDO-1, GSH metabolism, and HIF-1.

Issue 6: In addition, what are the major similarities and differences of cysteine homeostasis pathways between C. elegans and other systems (HIF dependency, transcription vs post-transcriptional control)? These questions could be better discussed and noted with novel findings of the current study that are likely C. elegans specific or broadly conserved.

Response 6: We have included a new section in the Discussion highlighting the nature of mammalian CDO1 regulation. We propose the hypothesis that a homologous pathway to the C. elegans RHY-1/CYSL-1/EGL9/HIF-1 pathway might operate in mammalian cells to sense high cysteine and induce CDO1 transcription. Importantly, all proteins in the C. elegans pathway have homologous counterparts in mammals. However, this hypothesis remains to be tested in mammalian systems.

Reviewer #3 (Public Review):

Major weaknesses of the paper include:

Issue 7: the over-reliance on genetic approaches.

Response 7: This is a fair critique. Our expertise is genetics. Our philosophy, which the reviewers may not share, is that there is no such thing as too much genetics!

Issue 8: the lack of novelty regarding prolyl hydroxylase-independent activities of EGL-9.

Response 8: We believe the primary novelty of our work is establishing the intersection between the H2Ssensing HIF-1 pathway and cysteine catabolism governed by cysteine dioxygenase. Our demonstration that cdo-1 regulation operates largely independent of VHL-1 and EGL-9 prolyl hydroxylation is a mechanistic detail of this regulation and not the critical new finding. Although, we believe it does suggest where pathway analyses should be directed in the future. We also believe that our homeostatic feedback model for the regulation of HIF-1 (and cdo-1) by cysteine-derived H2S is new and exciting and provides insight into the logic of why HIF-1 might respond to H2S and promote the activity of cdo-1. Our work suggests that one reason for this intersection of hif-1 and cdo-1 is to sense and maintain cysteine homeostasis when cysteine is in excess.

Issue 9: the lack of biochemical approaches to probe the underlying mechanism of the prolyl hydroxylaseindependent activity of EGL-9.

Response 9: While not the primary focus of our current manuscript, we agree that this is an exciting area of future research. To uncover the prolyl hydroxylase-independent activity of EGL-9, we agree that a combination of approaches will be required including, biochemical, structure-function, and genetic.

Major Issues We Feel the Authors Should Address:

Issue 10: One particularly glaring concern is that the authors really do not know the extent to which the prolyl hydroxylase activity is (or is not) impacted by the H487A mutation in egl-9(rae276). If there is a fair amount of enzymatic activity left in this mutant, then it complicates interpretation. The paper would be strengthened if the authors could show that the egl-9(rae276) eliminates most if not all prolyl hydroxylase activity. In addition, the authors may want to consider doing RNAi for egl-9 in the egl-9(rae276) mutant as a control, as this would support the claim that whatever non-hydroxylase activity EGL-9 may have is indeed the causative agent for the elevation of CDO-1::GFP. Without such experiments, readers are left with the nagging concern that this allele is simply a hypomorph for the single biochemical activity of EGL-9 (i.e., the prolyl hydroxylase activity) rather than the more interesting, hypothesized scenario that EGL-9 has multiple biochemical activities, only one of which is the prolyl hydroxylase activity.

Response 10: We have two lines of evidence that suggest the egl-9(rae276)-encoded H487A variant eliminates prolyl hydroxylase activity. First, Pan et al. 2007 (reference 57) demonstrate that when the equivalent histidine (H313) is mutated in human protein, that protein lacks detectible prolyl hydroxylase activity. Second, the phenotypic similarities caused by egl-9(rae276) and the vhl-1 null allele, ok161. Both alleles cause nearly identical activation of the Pcdo-1::GFP reporter transgene (Fig. 5C,D), and similarly impact the growth of the suox-1(gk738847) hypomorphic mutant (Table 1). This phenotypic overlap is highly relevant as the established role of VHL-1 is to recognize the hydroxyl mark conferred by the EGL-9 prolyl hydroxylase domain and promote the degradation of HIF-1. If EGL-9[H487A] had residual prolyl hydroxylase activity, we would expect the vhl-1(-) null mutant C. elegans to display more dramatic phenotypes than their egl-9(rae276) counterparts. This is not the case.

Issue 11: The authors observed that EGL-9 can inhibit HIF-1 and the expression of the HIF-1 target cdo-1 through a combination of activities that are (1) dependent on its prolyl hydroxylase activity (and subsequent VHL-1 activity that acts on the resulting hydroxylated prolines on HIF-1), and (2) independent of that activity. This is not a novel finding, as the authors themselves carefully note in their Discussion section, as this odd phenomenon has been observed for many HIF-1 target genes in multiple publications. While this manuscript adds to the description of this phenomenon, it does not really probe the underlying mechanism or shed light on how EGL-9 has these dual activities. This limits the overall impact and novelty of the paper.

Response 11: See response to Issues #8.

Issue 12: Cysteine dioxygenases like CDO-1 operate in an oxygen-dependent manner to generate sulfites from cysteine. CDO-1 activity is dependent upon availability of molecular oxygen; this is an unexpected characteristic of a HIF-1 target, as its very activation is dependent on low molecular oxygen. Authors neither address this in the text nor experimentally, and it seems a glaring omission.

Response 12: We agree this is an important point to raise within our manuscript. Although, despite its induction by HIF-1, there is no evidence that cdo-1 transcription is induced by hypoxia. In fact, in a genome wide transcriptomic study, cdo-1 was not found to be induced by hypoxia in C. elegans (Shen et al. 2005, reference 71).

We have newly commented on the use of molecular oxygen as a substrate by both EGL-9 and CDO-1 in our Discussion section. The mammalian oxygen-sensing prolyl hydroxylase (EGLN1) has been demonstrated to have high a Km value for O2 (high µM range). This likely allows EGLN1 to be poised to respond to small decreases in cellular oxygen from normal oxygen tensions. Clearly, CDO-1 also requires oxygen as a substrate, however the Km of CDO-1 for O2 is likely to be much lower, preventing sensitivity of the cysteine catabolism to physiological decreases in O2 availability. Although, to our knowledge, the CDO1 Km value for O2 has not been experimentally determined. We have added a new Discussion section where we address the conundrum about low oxygen inducing HIF-1 but oxygen being needed by CDO-1/CDO1.

Issue 13: The authors determined that the hypodermis is the site of the most prominent CDO-1::GFP expression, relevant to Figure 4. This claim would be strengthened if a negative control tissue, in the animal with the knockin allele, were shown. The hypodermal specific expression is a highlight of this paper, so it would make this article even stronger if they could further substantiate this claim.

Response 13: Our claim that the hypodermis is the critical site of cdo-1 function is based on; i) our hands on experience looking at Pcdo-1::GFP, Pcdo-1::CDO-1::GFP, CDO-1::GFP (encoded by cdo-1(rae273)) and our reporting of these expression patterns in multiple figures throughout the manuscript and ii) the functional rescue of cdo-1(-) phenotypes by a cdo-1 rescue construct expressed by a hypodermal-specific promoter (col10). We agree that providing negative control tissues would modestly improve the manuscript. However, we do not think that adding these controls will substantially alter the conclusions of the paper. Importantly, we acknowledge this limitation of our work with the sentence, “However, we cannot exclude the possibility that CDO-1 also acts in other cells and tissues as well.”

Minor issues to note:

Issue 14: Mutants for hif-1 and cysl-1 are sensitive to exogenous cysteine levels, yet loss of CDO-1 expression is not sufficient to explain this phenomenon, suggesting other targets of HIF-1 are involved. Given the findings the authors (and others) have had showing a role for RHY-1 in sulfur amino acid metabolism, shouldn't the authors consider testing rhy-1 mutants for sensitivity to exogenous cysteine?

Response 14: To test the hypothesis that rhy-1(-) C. elegans might be sensitive to supplemental cysteine, we cultured wild type and rhy-1(-) animals on 0, 100, and 1000µM supplemental cysteine. At 0 and 100µM supplemental cysteine, neither wild-type nor rhy-1(-) animals display any lethality suggesting rhy-1 is not required for survival in the face of excess cysteine (Fig. 3D,E). We also cultured these same strains on 1000µM supplemental cysteine, a concentration that is highly toxic to wild-type animals (100% lethality). rhy1(-) animals were resistant to 1000µM supplemental cysteine with a substantial fraction of the population surviving overnight exposure to this lethal dose of cysteine. Similarly, egl-9(-) mutant C. elegans were also resistant to 1000µM supplemental cysteine. We propose that loss of egl-9 or rhy-1 activates HIF-1-mediated transcription which is priming these mutants to cope with the lethal dose of cysteine. These data are now presented in Figure 3D-F and presented in the Results section.

Issue 15: The cysteine exposure assay was performed by incubating nematodes overnight in liquid M9 media containing OP50 culture. The liquid culture approach adds two complications: (1) the worms are arguably starving or at least undernourished compared to animals grown on NGM plates, and (2) the worms are probably mildly hypoxic in the liquid cultures, which complicates the interpretation.

Response 15: We agree that it is possible that animals growing overnight in liquid culture are undernourished and mildly hypoxic. However, we are confident in our data interpretation as all our experiments are appropriately controlled. Meaning, control and experimental groups were all grown under the same liquid culture conditions. Thus, these animals would all experience the same stressors that come with liquid culture. Importantly, we never make comparisons between groups that were grown under different culture conditions (i.e. solid media vs. liquid culture).

Issue 16: An easily addressable concern is the wording of one of the main conclusions: that cdo-1 transcription is independent of the canonical prolyl hydroxylase function of EGL-9 and is instead dependent on one of EGL-9's non-canonical, non-characterized functions. There are several points in which the wording suggests that CDO-1 toxicity is independent of EGL-9. In their defense, the authors try to avoid this by saying, "EGL-9 PHD," to indicate that it is the prolyl hydroxylase function of EGL-9 that is not required for CDO-1 toxicity. However, this becomes confusing because much of the field uses PHD and EGL-9/EGLN as interchangeable protein names. The authors need to be clear about when they are describing the prolyl hydroxylase activity of EGL-9 rather than other (hypothesized) activities of EGL-9 that are independent of the prolyl hydroxylase activity.

Response 16: We appreciate the reviewer alerting us to this practice within the field. To avoid confusion, we have removed the “PHD” abbreviation from our manuscript and explicitly referred to the “prolyl hydroxylase domain” where relevant.

Issue 17: The authors state in the text, "the egl-9; suox-1 double mutants are extremely sick and slow growing." We appreciate that their "health" assay, based on the exhaustion of food from the plate, is qualitative. We also appreciate that it is a functional measure of many factors that contribute to how fast a population of worms can grow, reproduce, and consume that lawn of food. However, unless they do a lifespan assay and/or measure developmental timing and specifically determine that the double mutant animals themselves are developing and/or growing more slowly, we do not think it is appropriate to use the words "slow growing" to describe the population. As they point out, the rate of consumption of food on the plate in their health assay is determined by a multitude and indeed a confluence of factors; the growth rate is one specific one that is commonly measured and has an established meaning.

Response 17: We see how the phrase ‘slow growing’ might imply a phenotype that we have not actually assessed with this assay. Therefore, we have removed all claims about “slow growth” of the strains presented in Table 1 and have highlighted the assay more overtly in the results section. For example; “While egl-9(-) and suox-1(gk738847) single mutant animals are healthy under standard culture conditions, the egl-9(-); suox1(gk738847) double mutant animals are extremely sick and require significantly more days to exhaust their E. coli food source under standard culture conditions (Table 1).”

Reviewer #1 (Recommendations For The Authors):

Issue 18: Relevance could be addressed further in the text.

Response 18: We have added additional context for our work in the Discussion section. Please see our response to Issues #5, 6, 12, and 24.

Issue 19: Better appreciation and integration of the manuscript's findings with published studies would be appropriate.

Response 19: We have added additional context for our work in the Discussion section. Please see our response to Issues #5, 6, 12, and 24.

Issue 20: It might be perhaps relevant to test whether cdo-1 is relevant for hypoxia resistance since it appears to be a key target for hif-1.

Response 20: We agree that this is an interesting future direction, however given that cdo-1 mRNA is not induced by hypoxia (Shen et al. 2005) we have not prioritized these experiments for the current manuscript.

Issue 21: "egl-9 inhibits cdo-1 transcription in a prolyl-hydroxylase and VHL-1-independent manner" should be tempered. vhl-1 mutants and egl-9 hydroxylase point mutant still have significant induction of the reporter.

Response 21: Thank you for identifying this oversight. We have modified the Figure 5 legend title to read, “egl9 inhibits cdo-1 transcription in a largely prolyl-hydroxylase and VHL-1-independent manner.”

Issue 22: Please use line numbers in the future for easier tracking of comments.

Response 22: We shall.

Issue 23: Abstract and elsewhere, "high cysteine activates...", should be rephrased to "high levels of cysteine".

Response 23: We have made this change throughout the manuscript.

Reviewer #3 (Recommendations For The Authors):

Issue 24: The authors discuss CDO1 in the context of tumorigenesis, as well as the potential regulation between cysteine and the hypoxia response pathway. Thus, I was surprised that there was no mention of the foundational Bill Kaelin paper (Briggs et al 2016) showing how the accumulation of cysteine is related to tumorigenesis, and that cysteine is a direct activator of EglN1. Puzzling that CDO1 is a tumor suppressor: you lose it, cysteine can accumulate and activate EglN1, causing HIF1 turnover. How do the authors reconcile their results with this paper? I was also surprised that there was no mention in the Discussion of the role of hydrogen sulfide, cysteine metabolism, and CTH and CBS in oxygen sensation in the carotid body given the role they play there. Seems important to discuss this issue.

Response 24: We have added new sections to our Discussion that consider the relationship between our work and Briggs et al. 2016 as well as mentioned the role of CTH and H2S in the mammalian carotid body.

Issue 25: The abstract has a variety of contradictory statements. For example, the authors state that "HIF-1mediated induction of cdo-1 functions largely independent of EGL-9," but then go on to conclude in the final sentence that cysteine stimulates H2S production, which then activates EGL-9 signaling, which then increases HIF-1-mediated transcription of cdo-1. A quick reading of the abstract leaves the reader uncertain whether EGL-9 is or is not involved in this regulation of cdo-1 expression. In addition, the conclusion sentence implies that activation of the EGL-9 pathway increases HIF-1-mediated transcription, yet it is well established that EGL-9 is an inhibitor of HIF-1. The abstract fails to deliver a clear summary of the paper's conclusions. Perhaps consider this alternative (changes in capital letters):

The amino acid cysteine is critical for many aspects of life, yet excess cysteine is toxic. Therefore, animals require pathways to maintain cysteine homeostasis. In mammals, high cysteine activates cysteine dioxygenase, a key enzyme in cysteine catabolism. The mechanism by which cysteine dioxygenase is regulated remains largely unknown. We discovered that C. elegans cysteine dioxygenase (cdo-1) is transcriptionally activated by high cysteine and the hypoxia inducible transcription factor (hif-1). hif-1- dependent activation of cdo-1 occurs downstream of an H2S-sensing pathway that includes rhy-1, cysl-1, and egl-9. cdo-1 transcription is primarily activated in the hypodermis where it is sufficient to drive sulfur amino acid metabolism. EGL-9 and HIF-1 are core members of the cellular hypoxia response. However, we demonstrate that the mechanism of HIF-1-mediated induction of cdo-1 IS largely independent of EGL-9 prolyl hydroxylASE ACTIVITY and the von Hippel-Lindau E3 ubiquitin ligase. We propose that the REGULATION OF cdo-1 BY HIF-1 reveals a negative feedback loop for maintaining cysteine homeostasis. High cysteine stimulates the production of an H2S signal. H2S then ACTS THROUGH the rhy-1/cysl-1/egl-9 signaling pathway DISTINCTLY FROM THEIR ROLE IN HYPOXIA RESPONSE TO INCREASE HIF-1-mediated transcription of cdo-1, promoting degradation of cysteine via CDO-1.

Response 25: We agree that the abstract could be clearer. We believe this concern stems from the fact that we did not discuss our initial screen in the abstract. Thus, we failed to establish a role for egl-9 in the regulation of cdo-1. To remedy this, we have modified the abstract as suggested by the reviewer and added additional context. We believe that these changes improve the clarity of the Abstract substantially.

Issue 26: An easily addressable concern involves the "dark" microscopy controls showing lack of fluorescence from a nematode. In these dark negative control micrographs, the authors should draw dotted outlines around where the worms are or include a brightfield image next to the fluorescence image. On a computer screen, it is in fact possible to make out the worms. Yet, when printed out, the reader must assume there are worms in the dark images. Additionally, we realize that adjusting fluorescence so that wild-type CDO-1 expression can be seen will result in oversaturation of the egl-9 and rhy-1; cdo-1 doubles; however, this would be a useful figure to add into the supplement to both provide a normal reference of CDO-1 low-level expression and a demonstration of just how bright it is in the mutant backgrounds. It would also be useful for you to please report your exposure settings for purposes of reproducibility.

Response 26: As suggested, we have added dotted lines around the location of the C. elegans animals in all images where GFP expression is low or basal. We have also reported the exposure times for each image in the appropriate figure legends.

Issue 27: This title is quite generic and doesn't even mention the main players (CDO-1 and sulfite metabolism).

Response 27: We have updated our title to call attention to cysteine dioxygenase. The improved title is: “Hypoxia-inducible factor induces cysteine dioxygenase and promotes cysteine homeostasis in Caenorhabditis elegans”

Issue 28: The authors mention two disorders in which CDO-1 plays a pathogenic role: MoCD and ISOD. We recommend switching the order in which the authors mention these, as the remainder of the paragraph is about MoCD. Also, they should write out the number "2" in the first sentence of that paragraph.

Response 28: We have made the suggested changes.

Issue 29: The authors state in the main text, "...to ubiquitinate HIF-1, targeting it for degradation by the proteosome." Here, they should refer to the pathway in Figure 5a.

Response 29: We have made the suggested change.

Issue 30: The authors state in the main text, "Elements of the HIF-1 pathway have emerged..." which is vague and confusingly worded. Change to, "Members of the HIF-1 pathway and its targets have emerged from C. elegans genetic studies."

Response 30: We have made the suggested change.

Issue 31: Clarify in the figure legends that supplemental cysteine did not affect the mortality of worms that were imaged.

Response 31: We have added this note to Figure 3A and Figure S3A.

Issue 32: Figure 1b. "the cdo-1 promoter is shown..." Add: "as a straight line" to the end of this phrase.

Response 32: We have made the suggested change.

Issue 33: The authors should consider changing the red text in Figure 1 to magenta, which tends to be more readable for people who have limited color vision.

Response 33: We have adjusted the colors in Figure 1 as suggested.

Issue 34: Figure 2, legend title. Consider changing "hif-1" to "HIF-1," as well as rhy-1, cysl-1, and egl-9. In this case, they are talking about proteins, not mutants or genes. This will make the paper easier to follow for readers who lack a C. elegans background.

Response 34: We have made the suggested change.

Issue 35: Figure 5, caption text. "...indicates weak similarity." Add, "amongst species compared."

Response 35: We have made the suggested change.

Issue 36: It is starting to become a standard for showing the datapoints in bar graphs. Although this is done in many graphs in the paper, it should also be done for Figure S1 and Figure 4C.

Response 36: We have made the suggested change.

Issue 37: An extensive ChIP-seq and RNA-seq analysis of C. elegans HIF-1 was recently published (Vora et al, 2022), which the authors should reference in support of the regulation of CDO-1 transcription by HIF-1 in their description of published expression studies of the pathway (Results section, page 4). Indeed, Vora et al were key generators of the ChIP-seq data cited in Warnhoff et al but not included as authors in the ModERN/ModENCODE publication: their contributions were published separately in Vora et al and should be acknowledged equivalently.

Response 37: We appreciate the reviewer pointing this detail out and we have added the correct citation as indicated.

Version 2

Abstract

Abstract

Dedicated genetic pathways regulate cysteine homeostasis. For example, high levels of cysteine activate cysteine dioxygenase, a key enzyme in cysteine catabolism in most animal and many fungal species. The mechanism by which cysteine dioxygenase is regulated is largely unknown. In an unbiased genetic screen for mutations that activate cysteine dioxygenase (cdo-1) in the nematode C. elegans, we isolated loss-of-function mutations in rhy-1 and egl-9, which encode proteins that negatively regulate the stability or activity of the oxygen-sensing hypoxia inducible transcription factor (hif-1). EGL-9 and HIF-1 are core members of the conserved eukaryotic hypoxia response. However, we demonstrate that the mechanism of HIF-1-mediated induction of cdo-1 is largely independent of EGL-9 prolyl hydroxylase activity and the von Hippel-Lindau E3 ubiquitin ligase, the classical hypoxia signaling pathway components. We demonstrate that C. elegans cdo-1 is transcriptionally activated by high levels of cysteine and hif-1. hif-1-dependent activation of cdo-1 occurs downstream of an H2S-sensing pathway that includes rhy-1, cysl-1, and egl-9. cdo-1 transcription is primarily activated in the hypodermis where it is also sufficient to drive sulfur amino acid metabolism. Thus, the regulation of cdo-1 by hif-1 reveals a negative feedback loop that maintains cysteine homeostasis. High levels of cysteine stimulate the production of an H2S signal. H2S then acts through the rhy-1/cysl-1/egl-9 signaling pathway to increase HIF-1-mediated transcription of cdo-1, promoting degradation of cysteine via CDO-1.

Introduction

Cysteine is a sulfur-containing amino acid that mediates many oxidation/reduction reactions of proteins, is the redox center of the abundant antioxidant tripeptide glutathione which also serves as a major cysteine reserve, and is essential for iron-sulfur cluster assembly in the mitochondrion (1, 2). Cysteine residues in many proteins are in close proximity in the primary or folded protein sequence and are oxidized in the endoplasmic reticulum to form intra- and interprotein disulfide linkages, most commonly in secreted proteins which mediate intercellular signaling and defense (3–5). In many enzymes, the reactivity of the cysteine sulfur is key for the coordination of metals such as zinc or iron, which support protein structure and catalytic activity (6–8). Cysteine is also a key source of hydrogen sulfide (H2S), a volatile signaling molecule (9). While cysteine has these essential functions, excess cysteine is also toxic. High levels of cysteine impair mitochondrial respiration by disrupting iron homeostasis (10), acts as a neural excitotoxin (11), and promotes the formation of toxic levels of hydrogen sulfide gas (9, 12, 13). Given this balance between essential and toxic, cysteine homeostasis is key for the health of cells and organisms.

CDO1-mediated oxidation is the primary pathway of cysteine catabolism when sulfur amino acid (methionine or cysteine) availability is normal or high (14). The dipeptide cystathionine is a key intermediate in this pathway. Cystathionine is catabolized by cystathionase (CTH-2 in C. elegans, CTH in mammals) producing cysteine and α-ketobutyrate. Cysteine is further oxidized using dissolved atmospheric dioxygen to cysteinesulfinate by cysteine dioxygenase (CDO-1 in C. elegans, CDO1 in mammals) (Fig. 1A) (15). The further oxidation of cysteinesulfinate downstream of CDO-1 generates highly toxic sulfites that are normally oxidized to more benign sulfate by sulfite oxidase (Fig. 1A).

egl-9 and rhy-1 inhibit cdo-1 transcription.

A) Pathway for sulfur amino acid metabolism beginning with methionine. We highlight the roles of cystathionase (CTH-2/CTH), cysteine dioxygenase (CDO-1/CDO1), and the Moco-requiring sulfite oxidase enzyme (SUOX-1/SUOX). C. elegans enzymes (magenta) and their human homologs (green) are displayed. B) Pcdo-1::GFP promoter fusion (upper) and Pcdo-1::CDO-1::GFP C-terminal protein fusion (lower) transgenes used in this work are displayed. Boxes indicate exons, connecting lines indicate introns. The cdo-1 promoter is shown as a straight line. C) egl-9a and rhy-1 gene structures. Boxes indicate exons and connecting lines are introns. Colored annotations indicate mutations generated or used in our work. Magenta; chemically-induced mutations that activated Pcdo-1::CDO-1::GFP fusion protein. Blue; reference null alleles isolated independent of our work. Green; CRISPR/Cas9-generated mutation that inactivates the prolyl hydroxylase domain of EGL-9. D) Expression of Pcdo-1::GFP transgene is displayed for wild-type, egl-9(sa307), and rhy-1(ok1402) C. elegans animals at the L4 stage. Scale bar is 250μm. White dotted line outlines animals with basal GFP expression. For GFP imaging, exposure time was 100ms. E) Quantification of GFP expression displayed in Fig. 1D. Individual datapoints are shown (circles) as are the mean and standard deviation (red lines). n is 5 individuals per genotype. Data are normalized so that wild-type expression of Pcdo-1::GFP is 1 arbitrary unit (a.u.). ****, p<0.0001, ordinary one-way ANOVA with Dunnett’s post hoc analysis.

As a critical player in cysteine homeostasis, CDO1 is a highly regulated enzyme: the activity and abundance of CDO1 increase dramatically in cells and animals fed excess cysteine and methionine (16). CDO1 levels are governed both transcriptionally and post-translationally via proteasomal degradation (17–21). However, the molecular players that sense high levels of cysteine and promote CDO1 activation remain incompletely defined.

Understanding these regulatory mechanisms is an important goal as CDO1 dysfunction is implicated in disease. CDO1 is a tumor suppressor whose activity is silenced in diverse cancers via promoter methylation, a potential biomarker of tumor grade and progression (22–24). Decreased CDO1 activity may support tumor cell growth by reducing reactive oxygen species and decreasing drug susceptibility (25–28).

Using the nematode C. elegans, we have previously shown that CDO-1 is a key player in the pathophysiology of two fatal inborn errors of metabolism; isolated sulfite oxidase deficiency (ISOD) and molybdenum cofactor deficiency (MoCD) (29, 30). Molybdenum cofactor (Moco) is an essential 520 Dalton prosthetic group synthesized from GTP by a conserved multistep biosynthetic pathway that is present in about 2/3 of bacterial genomes and nearly all eukaryotic genomes (31, 32). C. elegans can either retrieve Moco synthesized by the bacteria it consumes or can synthesize Moco de novo using its own Moco biosynthetic pathway (33). C. elegans strains carrying mutations in genes encoding Moco biosynthetic enzymes (moc) and feeding on wild-type E. coli develop normally. Yet, these same moc-mutant C. elegans when fed on Moco-deficient E. coli as their sole nutritional source are inviable, arresting development at an early larval stage. A saturated genetic selection for mutations that suppress this inviability identified multiple independent mutations in cth-2 or cdo-1, genes which encode enzymes in the cysteine biosynthetic and degradation pathway. The toxic sulfites produced downstream of CDO-1 are normally oxidized to more benign sulfate by Moco-requiring sulfite oxidase (SUOX-1 in C. elegans, SUOX in mammals) an essential enzyme in C. elegans and humans (Fig. 1A) (30, 33). Therefore, loss of cdo-1 or cth-2 suppresses the lethality caused by both Moco and sulfite oxidase deficiencies in C. elegans by preventing the production of toxic sulfites (33–35). Thus, understanding the fundamental mechanisms that govern the levels and activity of CDO-1 is critical to generating new therapeutic hypotheses to treat these diseases.

To define genes that regulate cdo-1 levels and activity, we performed an unbiased genetic screen in the nematode C. elegans to identify mutations that increase the expression or abundance of a Pcdo-1::CDO-1::GFP reporter transgene. We identified multiple independent loss-of-function mutations in two genes, egl-9 and rhy-1, that dramatically increase expression of this Pcdo-1::CDO-1::GFP transgene. These enzymes act in the hypoxia and H2S-sensing pathway, and we demonstrate that the conserved hypoxia-inducible transcription factor (HIF-1) activates cdo-1 transcription in this pathway (36–38). We further show that high levels of cysteine promote cdo-1 transcription, and that hif-1 and cysl-1 (another component of the H2S-sensing pathway) are required for viability under high cysteine conditions. We demonstrate that transcriptional activation of cdo-1 via HIF-1 promotes CDO-1 activity and establish the C. elegans hypodermis as a key tissue of CDO-1 activation and function. Unexpectedly, we find that cdo-1 regulation is governed by a HIF-1 pathway largely independent of EGL-9 prolyl hydroxylase activity and von Hippel-Lindau (VHL-1), the canonical O2 -sensing pathway (39, 40). These data establish a new connection between the HIF-1/H2S-sensing pathway and sulfur amino acid catabolism governed by CDO-1.

Results

egl-9 and rhy-1 negatively regulate cdo-1 transcription

To identify regulators of CDO-1 expression or activity, we engineered a transgene expressing a C-terminal green fluorescent protein (GFP) fusion to the full-length CDO-1 protein driven by the native cdo-1 promoter (Pcdo-1::CDO-1::GFP, Fig. 1B). Transgenic animals were generated by integrating the Pcdo-1::CDO-1::GFP fusion protein into the C. elegans genome (41). The Pcdo-1::CDO-1::GFP fusion protein was functional and rescued a cdo-1 loss of function mutation: the Pcdo-1::CDO-1::GFP fusion protein reverses the suppression of Moco-deficient lethality caused by cdo-1 loss of function (Fig. S1). The reanimation of Moco-deficient lethality by the transgene depends on CDO-1 enzymatic activity because a transgene expressing an active-site mutant Pcdo-1::CDO-1[C85Y]::GFP does not rescue the cdo-1 mutant phenotype (Fig. S1) (33, 42). Thus, the Pcdo-1::CDO-1::GFP transgenic fusion protein is functional, suggesting that its expression pattern reflects endogenous protein expression, localization, and levels.

We performed a mutagenesis screen to identify genes that control the expression or accumulation of CDO-1 protein. Specifically, we performed an EMS chemical mutagenesis of C. elegans and screened in the F2 generation, after newly induced random mutations were allowed to become homozygous, for mutations that caused increased GFP accumulation by the Pcdo-1::CDO-1::GFP reporter transgene (43). Using whole-genome sequencing, we determined that two independently isolated mutants carried distinct mutations in egl-9 and four other independently isolated mutants had unique mutations in rhy-1 (Fig. 1C). The presence of multiple independent alleles suggests these mutations in egl-9 or rhy-1 are causative for the increased Pcdo-1::CDO-1::GFP expression or accumulation observed. egl-9 encodes the O2 -sensing prolyl hydroxylase and orthologue of mammalian EglN1. rhy-1 encodes the regulator of hypoxia inducible transcription factor, an enzyme with homology to membrane-bound O-acyltransferases (36, 39). The genetic screen produced nonsense alleles of both egl-9 and rhy-1 suggesting the increased Pcdo-1::CDO-1::GFP expression or accumulation is caused by loss of egl-9 or rhy-1 function (Fig. 1C). Inactivation of egl-9 or rhy-1 activates a transcriptional program mediated by the hypoxia inducible transcription factor, HIF-1 (36, 39). To determine if egl-9 or rhy-1 regulate cdo-1 transcription, we engineered a separate reporter construct where only GFP (rather than the full length CDO-1::GFP fusion protein) is transcribed by the cdo-1 promoter (Pcdo-1::GFP, Fig. 1B). This Pcdo-1::GFP transgene was introduced into strains with independently isolated egl-9(sa307) and rhy-1(ok1402) null reference alleles (36, 44). The egl-9(sa307) and rhy-1(ok1402) mutations dramatically induce expression of GFP driven by the Pcdo-1::GFP transcriptional reporter transgene (Fig. 1D,E). The activation of cdo-1 transcription by independently isolated egl-9 or rhy-1 mutations demonstrates that the egl-9 and rhy-1 mutations isolated in our screen for the induction or accumulation of Pcdo-1::CDO-1::GFP are the causative genetic lesions. Furthermore, these data demonstrate that egl-9 and rhy-1 are necessary for the normal transcriptional repression of cdo-1.

HIF-1 directly activates cdo-1 transcription downstream of the rhy-1, cysl-1, egl-9 genetic pathway

rhy-1 and egl-9 act in a pathway that regulates the abundance and activity of the HIF-1 transcription factor (36, 38). The activity of rhy-1 is most upstream in the pathway and negatively regulates the activity of cysl-1, which encodes a cysteine synthase-like enzyme of probable algal origin (45). CYSL-1 directly binds to and inhibits EGL-9 in an H2S-modulated manner (38). EGL-9 uses molecular oxygen as well as an α-ketoglutarate cofactor to directly inhibit HIF-1 via prolyl hydroxylation, which recruits the VHL-1 ubiquitin ligase to ubiquitinate HIF-1, targeting it for degradation by the proteasome (Fig. 5A) (46–48). Given that loss-of-function mutations in rhy-1 or egl-9 activate cdo-1 transcription, we tested if cdo-1 transcription is activated by HIF-1 as an output of this hypoxia/H2S-sensing pathway. We performed epistasis studies using null mutations that inhibit the activity of hif-1 (cysl-1(ok762) and hif-1(ia4)) or activate hif-1 (rhy-1(ok1402) and egl-9(sa307)). The induction of Pcdo-1::GFP by egl-9 inactivation was dependent upon the activity of hif-1, but not on the activity of cysl-1 (Fig. 2A,B). In contrast, induction of Pcdo-1::GFP by rhy-1 inactivation was dependent upon the activity of both hif-1 and cysl-1 (Fig. 2A,C). These results reveal a genetic pathway whereby rhy-1, cysl-1, and egl-9 function in a negative-regulatory cascade to control the activity of HIF-1 which transcriptionally activates cdo-1. Our epistasis studies of cdo-1 transcriptional regulation by HIF-1 align well with previous analyses of this genetic pathway in the context of transcription of cysl-2 (a paralog of cysl-1) and the “O2-ON response” (38).

cdo-1 transcription is activated by HIF-1 downstream of RHY-1, CYSL-1, and EGL-9.

A) Expression of the Pcdo-1::GFP transgene is displayed for wild-type, egl-9(sa307), egl-9(sa307) hif-1(ia4) double mutant, egl-9(sa307); cysl-1(ok762) double mutant, rhy-1(ok1402), rhy-1(ok1402); hif-1(ia4) double mutant, and rhy-1(ok1402); cysl-1(ok762) double mutant C. elegans animals at the L4 stage of development. Scale bar is 250μm. White dotted line outlines animals with basal GFP expression. For GFP imaging, exposure time was 100ms. B,C) Quantification of the data displayed in Fig. 2A. Individual datapoints are shown (circles) as are the mean and standard deviation (red lines). n is 5 individuals per genotype. Data are normalized so that wild-type expression of Pcdo-1::GFP is 1 arbitrary unit (a.u.). *, p<0.05, ****, p<0.0001, ordinary one-way ANOVA with Dunnett’s post hoc analysis. Note, wild-type, egl-9(-), and rhy-1(-) images in panel A and quantification of Pcdo-1::GFP in panels B and C are identical to the data presented in Fig. 1D,E. They are re-displayed here to allow for clear comparisons to the double mutant strains of interest.

To demonstrate that HIF-1 activates transcription of endogenous cdo-1, we explored published RNA-sequencing data of wild-type, egl-9(-), and egl-9(-) hif-1(-) mutant animals (49). egl-9(-) mutant C. elegans display an 8-fold increase in cdo-1 mRNA compared to wild type. This induction was dependent on hif-1; a hif-1(-) mutation completely suppressed the induction of cdo-1 mRNA caused by an egl-9(-) mutation (49). These RNA-seq data confirm our findings using the Pcdo-1::GFP transcriptional reporter that HIF-1 is a transcriptional activator of cdo-1. ChIP-seq data of HIF-1 performed by the modERN project show that HIF-1 directly binds the cdo-1 promoter (peak from −1,165 to −714 base pairs 5’ to the cdo-1 ATG start codon) (50, 51). This HIF-1 binding site contains 3 copies of the HIF-binding motif (5’-RCGTG-3’) (52). Thus, cdo-1 is a downstream effector of HIF-1 and is likely a direct transcriptional target of HIF-1.

High levels of cysteine promote cdo-1 transcription and cause lethality in cysl-1 and hif-1 mutant animals

Mammalian CDO1 levels and activity are highly induced by dietary cysteine (14–16). To determine if this homeostatic response is conserved in C. elegans, we exposed transgenic C. elegans carrying the Pcdo-1::GFP transcriptional reporter to high supplemental cysteine. Like our egl-9 and rhy-1 loss-of-function mutations, high levels of cysteine promoted cdo-1 transcription (Fig. 3A-C). Despite the significant 3.6-fold induction of Pcdo-1::GFP caused by 100μM supplemental cysteine, we note that this induction is not as dramatic as the induction caused by null mutations in egl-9 or rhy-1. This perhaps reflects the animals’ ability to buffer environmental cysteine which is bypassed by genetic intervention.

High levels of cysteine activate cdo-1 transcription and are lethal to hif-1 and cysl-1 mutant animals.

A) Expression of the Pcdo-1::GFP transgene is displayed for wild-type, egl-9(sa307), and rhy-1(ok1402) young-adult C. elegans exposed to 0 or 100μM supplemental cysteine. Scale bar is 250μm. White dotted line exposure time was 100ms. Supplemental cysteine did not impact the mortality of the animals being imaged. B) Quantification of the data displayed in Fig. 3A. Individual datapoints are shown (circles) as are the mean and standard deviation (black lines). n is 6 or 7 individuals per genotype. Data are normalized so that expression of Pcdo-1::GFP in wild-type C. elegans exposed to 0μM supplemental cysteine is equal to 1 arbitrary unit (a.u.). *, p<0.05, ****, p<0.0001, multiple unpaired t test with Welch’s correction. C) Quantification of the Pcdo-1::GFP expression is displayed for wild-type young-adult C. elegans exposed to 0, 50, 100, 250 or 500μM supplemental cysteine. Mean and standard deviation are displayed. n is 6 or 7 individuals per concentration of supplemental cysteine. D-F) The percentage of animals that survive overnight exposure to D) 0, E) 100, or F) 1,000μM supplemental cysteine. Individual datapoints (circles) represent biological replicates. 3 or 4 biological replicates were performed for each experiment and the total individuals scored amongst all replicates is displayed (n). *, p<0.05, ****, p<0.0001, ordinary one-way ANOVA with Dunnett’s post hoc analysis. ns indicates no significant difference was identified.

We hypothesized that cysteine might activate cdo-1 transcription through the RHY-1/CYSL-1/EGL-9/HIF-1 pathway. In this pathway, cysl-1 and hif-1 act to promote cdo-1 transcription. Thus, we sought to test whether cysl-1 or hif-1 were necessary for the induction of Pcdo-1::GFP by high levels of cysteine. However, this experiment was not possible as we observed 100% lethality in cysl-1(-) and hif-1(-) mutant animals exposed to 100μM supplemental cysteine, a cysteine concentration at which wild-type animals are healthy (Fig. 3E). Importantly, wild-type, cysl-1(-) and hif-1(-) animals were all healthy under control conditions without supplemental cysteine (Fig. 3D). While this phenotype limits our ability to establish the role of cysl-1 and hif-1 in the induction of cdo-1 by high levels of cysteine, these data demonstrate that cysl-1 and hif-1 are necessary for survival under high cysteine conditions. Given the requirement of hif-1 for survival in high levels of cysteine, we hypothesized that mutations in egl-9 or rhy-1 that activate hif-1 might promote cysteine resistance. Indeed, we found that egl-9(-) and rhy-1(-) mutant C. elegans were partially viable when exposed to 1,000μM supplemental cysteine, a concentration that causes 100% lethality in wild-type animals (Fig. 3F). Thus, egl-9 and rhy-1 are negative regulators cysteine tolerance. Taken together, these data demonstrate a critical physiological role for the RHY-1/CYSL-1/EGL-9/HIF-1 pathway in promoting cysteine homeostasis.

To further test the interaction between high levels of cysteine and the RHY-1/CYSL-1/EGL-9/HIF-1 pathway, we exposed egl-9(-); Pcdo-1::GFP mutant animals to control or high levels of supplemental cysteine. We reasoned that if cysteine and egl-9 loss of function promote Pcdo-1::GFP accumulation in the same pathway, then their effects should not be additive. Consistent with this hypothesis, we saw no difference in Pcdo-1::GFP expression in egl-9(-) mutant animals exposed to 0 or 100μM supplemental cysteine (Fig. 3A,B). We also tested the ability of supplemental cysteine to further activate Pcdo-1::GFP expression in rhy-1(-) mutant animals. In contrast to our results with egl-9(-), we observed that high levels of cysteine caused a significant induction of Pcdo-1::GFP in the rhy-1(-) mutant background. These data suggest that cysteine acts in a pathway with egl-9 but operates in parallel to the function of rhy-1.

Given the established role of CDO-1 in cysteine catabolism, we tested whether cdo-1(-) mutants were also sensitive to high levels of cysteine. cdo-1(-) mutant animals were not sensitive to high levels of supplemental cysteine compared to the wild type (Fig. 3D-F). Thus, cdo-1 is not necessary for survival under high cysteine conditions. This suggests the existence of alternate pathways that promote cysteine homeostasis. Given the role of the RHY-1/CYSL-1/EGL-9/HIF-1 pathway in promoting cysteine homeostasis, we propose that HIF-1 activates pathways (in addition to cdo-1) that promote survival under high cysteine conditions.

Activated CDO-1 accumulates and is functional in the hypodermis

We sought to identify the site of action of CDO-1. To observe CDO-1 localization, we used CRISPR/Cas9 to insert the GFP open reading frame into the endogenous cdo-1 locus, replacing the native cdo-1 stop codon. This cdo-1(rae273) allele encodes a C-terminal tagged “CDO-1::GFP” fusion protein from the native cdo-1 genomic locus (Fig. 4A). To determine if CDO-1::GFP was functional, we combined the CDO-1::GFP fusion protein with a null mutation in moc-1, a gene that is essential for C. elegans Moco biosynthesis (33). We then observed the growth of moc-1(-) CDO-1::GFP C. elegans on wild-type and Moco-deficient E. coli. The moc-1(-) mutant C. elegans expressing CDO-1::GFP from the native cdo-1 locus arrest during larval development when fed Moco-deficient E. coli, but not when fed Moco-producing E.coli (Fig. 4C). This lethality is caused by the CDO-1-mediated production of sulfites which are only toxic when C. elegans is Moco deficient, and demonstrates that the CDO-1::GFP fusion protein is functional (33). These data suggest that the CDO-1::GFP expression we observe is physiologically relevant.

Hypodermal CDO-1 accumulates in the cytoplasm when egl-9 or rhy-1 are inactive and is sufficient to promote sulfur amino acid metabolism.

A) Diagram of cdo-1(rae273), a CRISPR/Cas9-generated allele with GFP inserted into the cdo-1 gene, creating a functional C-terminal CDO-1::GFP fusion protein expressed from the native cdo-1 locus. B) Differential interference contrast (DIC) and fluorescence imaging are shown for wild-type, egl-9(sa307), and rhy-1(ok1402) C. elegans expressing CDO-1::GFP encoded by cdo-1(rae273). Scale bar is 10μm. For GFP imaging, exposure time was 200ms. An anterior segment of the Hyp7 hypodermal cell is displayed. CDO-1::GFP accumulates in the cytoplasm and is excluded from the nuclei. C) moc-1(ok366), moc-1(ok366) cdo-1(mg622), and moc-1(ok366) cdo-1(rae273) animals were cultured from synchronized L1 larvae for 72 hours on wild-type (black, Moco+) or ΔmoaA mutant (red, Moco-) E. coli. D) moc-1(ok366) cdo-1(mg622) double mutant animals expressing Pcol-10::CDO-1::GFP or Pcol-10::CDO-1[C85Y]::GFP transgenes were cultured for 48 hours on wild-type (black, Moco+) or ΔmoaA mutant (red, Moco-) E. coli. Two independently derived strains were tested for each transgene. For panels C and D, animal lengths were determined for each condition. Individual datapoints are shown (circles) as are the mean and standard deviation. Sample size (n) is 10 individuals for each experiment. ****, p<0.0001, multiple unpaired t test with Welch’s correction. ns indicates no significant difference was identified.

When wild-type animals expressing CDO-1::GFP were grown under standard culture conditions, we observed CDO-1::GFP expression in multiple tissues, including prominent expression in the hypodermis (Fig. 4B, Fig. S2). We tested if CDO-1::GFP fusion protein levels were affected by inactivation of egl-9 or rhy-1. We generated egl-9(-); CDO-1::GFP and rhy-1(-); CDO-1::GFP animals, and assayed expression of CDO-1::GFP. We found that CDO-1::GFP fusion protein levels, encoded by cdo-1(rae273), were increased by egl-9(-) or rhy-1(-) mutations (Fig. 4B, Fig. S2). This is consistent with our studies using Pcdo-1::CDO-1::GFP and Pcdo-1::GFP transgenes. Furthermore, high levels of cysteine also promoted accumulation of CDO-1::GFP (Fig. S3). In all scenarios, the hypodermis was the most prominent site of CDO-1::GFP accumulation. Specifically, we found that CDO-1::GFP was expressed in the cytoplasm of Hyp7, the major C. elegans hypodermal cell (Fig. 4B).

Based on the expression pattern of CDO-1::GFP encoded by cdo-1(rae273), we hypothesized that CDO-1 acts in the hypodermis to promote sulfur amino acid metabolism. To test this hypothesis, we engineered a cdo-1 rescue construct in which cdo-1 is expressed exclusively in the hypodermis under the control of a hypoderm-specific collagen (col-10) promoter (Pcol-10::CDO-1::GFP) (53). Collagens are expressed exclusively in the hypodermis of nematodes, with a periodic induction in phase with the almost diurnal molting cycle (54). Multiple independent transgenic C. elegans strains were generated by integrating the Pcol-10::CDO-1::GFP construct into the C. elegans genome and tested for rescue of the cdo-1(-) mutant suppression of Moco-deficient larval arrest (41). Thus, tissue-specific complementation of the cdo-1 mutation would regenerate a lethal arrest phenotype caused by Moco deficiency. We found that multiple independent transgenic strains of cdo-1(-) moc-1(-) double mutant animals carrying the Pcol-10::CDO-1::GFP transgene displayed a larval arrest phenotype when fed a Moco-deficient diet (Fig. 4D). These data demonstrated that hypodermal-specific expression of cdo-1 is sufficient to rescue the cdo-1(-) mutant suppression of Moco-deficient lethality. This rescue was dependent upon the enzymatic activity of CDO-1 as an active site variant of this transgene (Pcol-10::CDO-1[C85Y]::GFP) did not rescue the suppressed larval arrest of cdo-1(-) moc-1(-) double mutant animals fed Moco-deficient diets (Fig. 4D). Taken together, our analyses of CDO-1::GFP expression demonstrate that CDO-1 is expressed, and that expression is regulated, in multiple tissues, principal among them being the hypodermis and Hyp7 cell. Our tissue-specific rescue data demonstrate that hypodermal expression of cdo-1 is sufficient to promote cysteine catabolism and suggest that the hypodermis is a critical tissue for sulfur amino acid metabolism. However, we cannot exclude the possibility that CDO-1 also acts in other cells and tissues as well.

HIF-1 promotes CDO-1 activity downstream of the H2S-sensing pathway

We sought to determine the physiological impact of cdo-1 transcriptional activation by HIF-1. We reasoned mutations that activate HIF-1 and increase cdo-1 transcription may cause increased CDO-1 activity. CDO-1 sits at a critical metabolic node in the degradation of the sulfur amino acids cysteine and methionine (Fig 1A). A key byproduct of sulfur amino acid metabolism and CDO-1 is sulfite, a reactive toxin that is detoxified by the Moco-requiring sulfite oxidase (SUOX-1). Null mutations in suox-1 cause larval lethality. However, animals carrying the suox-1(gk738847) hypomorphic allele are healthy under standard culture conditions (33, 34). suox-1(gk738847) mutant animals display only 4% SUOX-1 activity compared to wild type and are exquisitely sensitive to sulfite stress (55). Thus, the suox-1(gk738847) mutation creates a sensitized genetic background to probe for increases in endogenous sulfite production. To test if increased cdo-1 transcription would impact the growth of suox-1-comprimised animals, we combined the egl-9 null mutation, which promotes HIF-1 activity and cdo-1 transcription, with the suox-1(gk738847) allele. While egl-9(-) and suox-1(gk738847) single mutant animals are healthy under standard culture conditions, the egl-9(-); suox-1(gk738847) double mutant animals are extremely sick and require significantly more days to exhaust their E. coli food source under standard culture conditions (Table 1). These data establish a synthetic genetic interaction between these loci. To determine the role of sulfur amino acid metabolism in the egl-9(-); suox-1(gk738847) synthetic sickness phenotype, we engineered egl-9(-); cdo-1(-) suox-1(gk738847) and cth-2(-); egl-9(-); suox-1(gk738847) triple mutant animals. The egl-9; suox-1 synthetic sickness phenotype was suppressed by inactivating mutations in cdo-1 or cth-2 which block the endogenous production of sulfite (Table 1). These data demonstrate that the deleterious activity of the egl-9(-) mutation in a suox-1(gk738847) background requires functional sulfur amino acid metabolism.

To determine the role of the H2S-sensing pathway in the synthetic sickness phenotype displayed by egl-9(-); suox-1(gk738847) double mutant animals, we introduced null alleles of cysl-1 or hif-1 into the egl-9(-); suox-1(gk738847) double mutant. The egl-9; suox-1 synthetic sickness phenotype was dependent upon hif-1 but not cysl-1 (Table 1). These results are consistent with our proposed genetic pathway and support the model that transcriptional activation of cdo-1 by HIF-1 causes increased CDO-1 activity and increased flux of sulfur amino acids through their catabolic pathway.

Loss of rhy-1 also strongly activates cdo-1 transcription. We hypothesized that rhy-1(-); suox-1(gk738847) double mutant animals would display a synthetic sickness phenotype like egl-9(-); suox-1(gk738847) double mutant animals. However, based upon their ability to exhaust their E. coli food source under standard culture conditions, rhy-1(-); suox-1(gk738847) double mutant animals were just as healthy as either rhy-1(-) or suox-1(gk738847) single mutant C. elegans (Table 1). These data suggest that increasing cdo-1 transcription alone is not sufficient to promote sulfite production via CDO-1. In addition to the role played by rhy-1 in the regulation of HIF-1 activity, rhy-1 itself is a transcriptional target of HIF-1. Loss of egl-9 activity induces rhy-1 mRNA ∼50-fold in a hif-1-dependent manner (49). Given this potent transcriptional activation, we wondered if rhy-1 might play an additional role downstream of HIF-1 in the regulation of sulfur amino acid metabolism and sulfite production. To test this hypothesis, we engineered rhy-1(-); egl-9(-); suox-1(gk738847) triple mutant animals. To our surprise, the rhy-1(-); egl-9(-); suox-1(gk738847) triple mutant animals were healthy, demonstrating that rhy-1 was necessary for the deleterious activity of the egl-9(-) mutation in a suox-1(gk738847) background (Table 1). These genetic data suggest a dual role for rhy-1 in the control of sulfur amino acid metabolism; first as a component of a regulatory cascade that controls the activity of HIF-1 and second as a functional downstream effector of HIF-1 that is required for sulfur amino acid metabolism.

This is not the first description of a rhy-1 role downstream of hif-1. Overexpression of a rhy-1-encoding transgene suppresses the lethality of a hif-1(-) mutant during HSS stress (56). These data establish RHY-1 as both a regulator and effector of HIF-1. How RHY-1, a predicted membrane-bound O-acyltransferase, molecularly executes these dual roles remains to be explored.

EGL-9 prolyl hydroxylase activity and VHL-1 are largely dispensable in the regulation of CDO-1

EGL-9 inhibits HIF-1 through its prolyl hydroxylase domain that hydroxylates HIF-1 proline 621 (Fig. 5A) (39). To evaluate the impact of the EGL-9 prolyl hydroxylase domain on the regulation of cdo-1, we generated a prolyl hydroxylase domain-inactive egl-9 mutation using CRISPR/Cas9. We engineered an egl-9 mutation that substitutes an alanine in place of histidine 487 (H487A) (Fig. 1C). Histidine 487 of EGL-9 is highly conserved and catalytically essential in the prolyl hydroxylase domain active site (Fig. 5B) (57, 58). To evaluate the impact of an inactive EGL-9 prolyl hydroxylase domain on the transcription of cdo-1, we engineered a C. elegans strain carrying the egl-9(H487A) mutation with the Pcdo-1::GFP transcriptional reporter. egl-9(H487A) caused a modest increase in Pcdo-1::GFP accumulation in the C. elegans intestine, suggesting that the EGL-9 prolyl hydroxylase domain is necessary to repress cdo-1 transcription in the intestine (Fig. 5C,D). However, the activation of Pcdo-1::GFP by the egl-9(H487A) mutation was markedly less when compared to Pcdo-1::GFP activation caused by an egl-9(-) null mutation (Fig. 5C,D). These data suggest that EGL-9 has a prolyl hydroxylase domain-independent activity that is responsible for repressing cdo-1 transcription.

egl-9 inhibits cdo-1 transcription in a largely prolyl-hydroxylase and VHL-1-independent manner.

A) The pathway of HIF-1 processing during normoxia is displayed. EGL-9 uses O2 as a substrate to hydroxylate (-OH) HIF-1 on specific proline residues. Prolyl hydroxylated HIF-1 is bound by VHL-1 which facilitates HIF-1 polyubiquitination and targets HIF-1 for degradation by the proteasome. B) Amino acid alignment of the EGL-9 prolyl hydroxylase domain from C. elegans, D. melanogaster, M. musculus, and H. sapiens. “*” indicate perfect amino acid conservation while “.” indicates weak similarity amongst species compared. Highlighted (red) is the catalytically essential histidine 487 residue in C. elegans. Alignment was performed using Clustal Omega (EMBL-EBI). C) Expression of Pcdo-1::GFP promoter fusion transgene is displayed for wild-type, egl-9(sa307, -), egl-9(rae276, H487A), and vhl-1(ok161) C. elegans animals at the L4 stage of development. Scale bar is 250μm. White dotted line outlines animals with basal GFP expression. For GFP imaging, exposure time was 500ms. D) Quantification of the data displayed in Fig. 5C. Individual datapoints are shown (circles) as are the mean and standard deviation (red lines). n is 6 individuals per genotype. Data are normalized so that wild-type expression of Pcdo-1::GFP is 1 arbitrary unit (a.u.). ***, p<0.001, ****, p<0.0001, ordinary one-way ANOVA with Tukey’s multiple comparisons test. ns indicates no significant difference was identified.

EGL-9 hydroxylates specific proline residues on HIF-1. These hydroxylated proline residues are recognized by the von Hippel-Lindau E3 ubiquitin ligase (VHL-1). VHL-1-mediated ubiquitination promotes degradation of HIF-1 by the proteasome (Fig. 5A) (40). Thus, the EGL-9 prolyl hydroxylase domain and VHL-1 act in a pathway to regulate HIF-1. To determine the role of VHL-1 in regulating cdo-1, we engineered a C. elegans strain carrying the vhl-1(-) null mutation with our Pcdo-1::GFP reporter. vhl-1 inactivation also caused a modest increase in Pcdo-1::GFP expression in the C. elegans intestine, suggesting that vhl-1 is necessary to repress cdo-1 transcription (Fig. 5C,D). However, activation of Pcdo-1::GFP by the vhl-1(-) mutation was less than activation caused by an egl-9(-) null mutation (Fig. 5C,D). These data suggest that EGL-9 has a VHL-1-independent activity that is responsible for repressing cdo-1 transcription.

To evaluate the impact of inactivating the EGL-9 prolyl hydroxylase domain or VHL-1 on cysteine metabolism, we again employed the suox-1(gk738847) hypomorphic mutation that sensitizes animals to increases in sulfite. We engineered egl-9(H487A); suox-1(gk738847) and vhl-1(-) suox-1(gk738847) double mutant animals and evaluated the health of those strains. In contrast to egl-9(-); suox-1(gk738847) double mutant animals which are extremely sick, egl-9(H487A); suox-1(gk738847) and vhl-1(-) suox-1(gk738847) double mutant animals are healthy (Table 1). These genetic data suggest that neither the EGL-9 prolyl hydroxylase domain nor VHL-1 are necessary to repress cysteine catabolism/sulfite production. However, it is plausible that the egl-9(H487) or vhl-1(-) mutations modestly activate cysteine metabolism, likely proportional to their activation of the Pcdo-1::GFP transgene (Fig. 5C,D), and that this activation is not sufficient to produce enough sulfites to negatively impact the growth of suox-1(gk738847) mutant animals.

Discussion

CDO-1 is a physiologically relevant effector of HIF-1

The hypoxia-inducible factor HIF-1 is a master regulator of the cellular response to hypoxia. It activates the transcription of many genes and pathways that are critical to maintain metabolic homeostasis in the face of low O2. For example, mammalian HIF1α induces the hematopoietic growth hormone erythropoietin, glucose transport and glycolysis, as well as lactate dehydrogenase (59–62). The nematode C. elegans encounters a range of O2 tensions in its natural habitat of rotting material: as microbial abundance increases, O2 levels decrease from atmospheric levels (∼21% O2). In fact, C. elegans prefers 5-12% O2, perhaps because hypoxia predicts abundant bacterial food sources (63). Members of the HIF-1 pathway and its targets have emerged from genetic studies of C. elegans (39). For instance, C. elegans HIF-1 promotes H2S homeostasis by inducing transcription of the mitochondrial sulfide quinone oxidoreductase (sqrd-1), detoxifying H2S (37).

We sought to define genes that regulate the levels and activity of cysteine dioxygenase (CDO-1), a critical regulator of cysteine homeostasis (14–16). Taking an unbiased genetic approach in C. elegans, we found that cdo-1 was highly regulated by HIF-1 downstream of a signaling pathway that includes rhy-1, cysl-1, and egl-9. We demonstrated that HIF-1 promotes cdo-1 transcription, accumulation of CDO-1 protein, and increased CDO-1 activity. Loss of rhy-1 or egl-9 activate hif-1 to in turn strongly induce the cdo-1 promoter. The pathway for activation of cdo-1 also requires cysl-1, which functions downstream of rhy-1 and upstream of egl-9. Based on ChIP-Seq studies, HIF-1 directly binds to the cdo-1 promoter (50). Interestingly, mammalian CDO1 is regulated at both the transcriptional and post-transcriptional level in response to high dietary cysteine (17–21). Our studies in the nematode C. elegans suggest that CDO1 transcription in mammals might be governed by HIF1α in response to changes in cellular cysteine. Importantly, all members of the RHY-1/CYSL-1/EGL-9 pathway in C. elegans have homologs encoded by mammalian genomes (38). Given the conservation of these proteins, future studies may show that similar cysteine and H2S-responsive signaling pathways operate in mammals.

We also show that the transcriptional activation of cdo-1 by HIF-1 promotes CDO-1 enzymatic activity. Genetic activation of cdo-1 by loss of egl-9 causes severe sickness in a mutant with reduced sulfite oxidase activity, an activity required to cope with the toxic sulfite produced via CDO-1. This synthetic sickness is dependent upon an intact HIF-1 signaling pathway and a functioning sulfur amino acid metabolism pathway, as the dramatic sickness of an egl-9; suox-1 double mutant is suppressed by loss of hif-1, cth-2, or cdo-1. These genetic results validate the intersection of HIF-1 and CDO-1 in converging biological regulatory pathways and encourage further exploration of this regulatory node. The potential physiological relevance of the connection between HIF-1 signaling and cysteine metabolism is discussed below.

It was initially surprising that the canonical hypoxia-sensing transcription factor, that is activated by relatively low O2 tensions, induces transcription of cdo-1, which encodes an oxidase that requires dissolved O2 to oxidize cysteine. Both EGL-9 and CDO-1 are dioxygenases, requiring O2 as a substrate to catalyze their chemical reactions. EGL-9 functions as an O2 sensor, using its O2 substrate to hydroxylate HIF-1 at relatively high O2 partial pressures, targeting it for degradation. EGL-9 is uniquely poised to sense deviations from the normally high physiological O2 concentrations given its high Km for O2 (64–68). While it is not known if mammalian CDO1 or C. elegans CDO-1 dioxygenases are active at lower O2 tensions than EGL-9, it is possible, even probable, that the set point for activation of HIF-1 by the failure of EGL-9 prolyl hydroxylation, is at a higher O2 tension than the Km of CDO-1 for oxidation of cysteine. In this way, CDO-1 could still coordinate cysteine catabolism while the cell is experiencing hypoxia and EGL-9 is unable to hydroxylate HIF-1.

Evidence for distinct pathways of HIF-1 activation by hypoxia and cysteine/H2S

The hypoxia-signaling pathway is defined by EGL-9-dependent prolyl hydroxylation of HIF-1. Hydroxylated HIF-1 is then targeted for degradation via VHL-1-mediated ubiquitination (39, 40, 52). However, multiple lines of evidence, reinforced by our work, demonstrate VHL-1- and prolyl hydroxylase-independent activity of EGL-9. egl-9(-) null mutant C. elegans accumulate HIF-1 protein and display increased transcription of many genes, including nhr-57, an established target of HIF-1 (36). In rescue experiments of an egl-9(-) null mutant, Shao et al. (2009) demonstrate that a wild-type egl-9 transgene restores normal HIF-1 protein levels and HIF-1 transcription. However, rescue experiments with a prolyl hydroxylase domain-inactive egl-9(H487A) transgene do not correct the accumulation of HIF-1 protein and only partially reduce the HIF-1 transcriptional output (58). Through our studies of cdo-1 transcription, we demonstrate that an egl-9(H487A) mutant incompletely activates HIF-1 transcription when compared to an egl-9(-) null mutation (Fig. 5C,D). Taken together, we conclude EGL-9 has activity independent of its prolyl hydroxylase domain, mirroring and supporting previous work (58).

In studies of the HIF-1-dependent Pnhr-57::GFP transcriptional reporter, Shen et al. (2006) observed that egl-9(-) null mutants promote HIF-1 transcription more than a vhl-1(-) mutant (36). We observe this same distinction between egl-9 and vhl-1 mutations with our Pcdo-1::GFP reporter (Fig. 5C,D). This difference in transcriptional activity is not explained by HIF-1 protein levels as HIF-1 protein accumulated equally in egl-9(-) and vhl-1(-) null mutants (36). This observation suggests that EGL-9 represses both HIF-1 levels and activity. This study also notes that vhl-1 represses HIF-1 transcription in the intestine while egl-9 acts in a wider array of tissues including the intestine and the hypodermis (36). Budde and Roth (2010) also demonstrate that loss of vhl-1 promotes HIF-1 transcription in the C. elegans intestine while total loss of egl-9 promotes HIF-1 transcription in multiple tissues including the intestine and the hypodermis (69). Our data expand upon these observations by demonstrating that loss of the EGL-9 prolyl hydroxylase domain promotes cdo-1 transcription in the intestine alone, mirroring the loss of vhl-1 (Fig. 5C). Budde and Roth (2010) additionally demonstrate that physiologically relevant stimuli also elicit a tissue-specific transcriptional response: hypoxia promotes HIF-1 transcription in the intestine while H2S promotes HIF-1 transcription in the hypodermis. Importantly, H2S promotes hypodermal HIF-1 transcription in a vhl-1(-) mutant, demonstrating a VHL-1-independent pathway for H2S activation of HIF-1 (69). We demonstrate that high levels of cysteine promote cdo-1 transcription in the hypodermis. Taken together with our data, these studies suggest two distinct pathways for activating HIF-1 transcription: i) a hypoxia-sensing pathway that is dependent upon vhl-1 and the EGL-9 prolyl hydroxylase domain and promotes HIF-1 activity in the intestine and ii) an H2S-sensing pathway that is independent of vhl-1 and the EGL-9 prolyl hydroxylase domain and promotes HIF-1 activity in the hypodermis.

The focus of CDO-1 expression and regulation of sulfite production in the hypoderm may be due to the demand in sulfur metabolism as well as oxygen-dependent hydroxylation of collagens during the C. elegans molting cycle. Many collagen genes are expressed before each larval molt and many collagen prolines are hydroxylated, like particular HIF-1 prolines, and their cysteines form disulfides during collagen assembly. The large demand for the amino acid cysteine in the many collagen genes expressed at high levels during a molt may challenge cysteine homeostasis in the hypodermis (54).

The genetic details of the H2S-sensing pathway were solidified through studies of rhy-1 in C. elegans. Ma et al. (2012) demonstrate that hif-1 repression via rhy-1 requires the activity of cysl-1, a gene encoding a cysteine synthase-like protein (38). They further demonstrate that high H2S promotes a physical interaction between CYSL-1 and EGL-9, resulting in the inactivation of EGL-9 and increased HIF-1 activity. Together, these studies suggest distinct genetic regulators of EGL-9/HIF-1 signaling: rhy-1 and cysl-1 govern the H2S-sensing pathway while vhl-1 mediates the hypoxia-sensing pathway. These pathways are distinct in their requirement for the EGL-9 prolyl hydroxylase domain.

A negative feedback loop senses high cysteine/H2S, promotes CDO-1 activity, and maintains cysteine homeostasis

Why would the RHY-1/CYSL-1/EGL-9/HIF-1 H2S-sensing pathway control the levels and activity of cysteine dioxygenase? We speculate this intersection facilitates a homeostatic pathway allowing C. elegans to sense and respond to cysteine level. We propose that H2S acts as a gaseous signaling molecule to promote cysteine catabolism. H2S activates HIF-1 in the hypodermis by promoting the CYSL-1-mediated inactivation of EGL-9 (38). We show that high levels of cysteine similarly induce cdo-1 transcription in the hypodermis. Our genetic data demonstrate that cdo-1 is induced by the same genetic pathway that senses H2S in C. elegans and CDO-1 acts in the hypodermis, the major site of H2S-induced transcription. Furthermore, H2S induces endogenous cdo-1 transcription >3-fold while cdo-1 mRNA levels do not change when C. elegans are exposed to hypoxia (70, 71). Thus, it is likely that H2S promotes cdo-1 transcription through RHY-1, CYSL-1, EGL-9, and HIF-1. H2S is a reasonable small molecule signal to alert cells to high levels of cysteine. Excess cysteine results in the production of H2S mediated by multiple enzymes including cystathionase (CTH), cystathionine β-synthase (CBS), and 3-mercaptopyruvate sulfurtransferase (MST) (9, 72). For example, CTH activity within the carotid body produces H2S that modifies the mammalian response to hypoxia (73). We speculate that excess cysteine in C. elegans promotes the enzymatic production of H2S which activates HIF-1 via the RHY-1/CYSL-1/EGL-9 signaling pathway. In our homeostatic model, H2S-activated HIF-1 would then induce cdo-1 transcription, promoting CDO-1 activity and the catabolism of the high-cysteine trigger (Fig. 6). Supporting this model, cysl-1(-) and hif-1(-) mutant C. elegans cannot survive in a high cysteine environment, demonstrating their central role in promoting cysteine homeostasis.

Model for the regulation of cysteine metabolism by HIF-1.

A) Proposed genetic pathway for the regulation of cdo-1. rhy-1, cysl-1, and egl-9 act in a negative-regulatory cascade to control activity of the HIF-1 transcription factor, which activates transcription of cdo-1. B) Under basal conditions (inactivated HIF-1), EGL-9 negatively regulates HIF-1 through 2 distinct pathways; one pathway is dependent upon O2, prolyl hydroxylation (Pro-OH), and VHL-1, while the second acts independently of these canonical factors. Under these conditions cdo-1 transcription is kept at basal levels and cysteine catabolism is not induced. C) During conditions where HIF-1 is activated (high H2S), CYSL-1 directly binds and inhibits EGL-9, preventing HIF-1 inactivation. Active HIF-1 binds the cdo-1 promoter, driving transcription and promoting CDO-1 protein accumulation. High CDO-1 levels promote the catabolism of cysteine leading to production of sulfites (SO32-) that are toxic during Moco or SUOX-1 deficiency. HIF-1-induced cysteine catabolism requires the activity of rhy-1.

Importantly, the human ortholog of EGL-9 (EglN1) has previously been implicated as a cysteine sensor in the context of triple negative breast cancer (TNBC) (74). This study determined that HIF1α accumulates in TNBC cells even during normoxia. Briggs et al. (2016) demonstrate that L-glutamate secretion via TNBC cells suppresses HIF1α prolyl hydroxylation, stabilizing HIF1α. L-glutamate secretion is mediated via the glutamate/cystine antiporter, xCT (75, 76). L-glutamate secretion inhibits xCT and a concomitant decrease in intracellular cystine/cysteine was observed. The authors propose that low intracellular cystine/cysteine produces oxidizing conditions that oxidize specific EglN1 cysteine residues, inhibiting the activity of EglN1. Thus, Briggs et al. (2016) propose EglN1 as a cysteine sensor whose activity is promoted by cysteine.

Our work in C. elegans also strongly suggests a role for EGL-9 in sensing and responding to cysteine. We show that high levels of cysteine promote transcription of the HIF-1 target gene cdo-1, and that this induction is not additive with a null mutation in egl-9, suggesting cysteine and egl-9 act in a pathway. Therefore, we propose that in C. elegans high levels of cysteine inhibit the activity of EGL-9, the opposite effect observed in TNBC cells. Furthermore, our genetic studies demonstrate that regulation of cdo-1 transcription occurs largely independent of the EGL-9 prolyl hydroxylase domain and VHL-1, while the mechanism proposed by Briggs et al. (2016) suggests that the stabilization of HIF1α by L-glutamate is correlated with increased prolyl hydroxylation. Despite these differences, it seems likely that the roles for C. elegans EGL-9 and human EglN1 in sensing cysteine are connected. However, additional studies are required to determine if there is an evolutionary relationship between these cysteine-sensing mechanisms.

Members of the H2S-sensing pathway have also been implicated in the C. elegans response to infection (44, 77). Many secreted proteins that mediate intercellular signaling or innate immune responses to pathogens are cysteine-rich and form disulfide bonds in the endoplasmic reticulum before protein secretion (78, 79). Levels of free cysteine may fall during the massive inductions of cysteine-rich secreted proteins in development or immune defense, as well as during the synthesis of collagens in the periodic molting cycle (54, 80). The oxidation of so many cysteines to disulfides in the endoplasmic reticulum might locally lower O2 levels preventing EGL-9 hydroxylation of HIF-1. Thus, the intersection between HIF-1 and cysteine homeostasis we have uncovered may contribute to a regulatory axis in cell-cell signaling during development and in immune function.

Methods

General methods and strains

C. elegans were cultured using established protocols (43). Briefly, animals were cultured at 20°C on nematode growth media (NGM) seeded with wild-type E. coli (OP50). The wild-type strain of C. elegans was Bristol N2. Additional E. coli strains used in this work were BW25113 (Wild type, Moco+) and JW0764-2 (ΔmoaA753::kan, Moco-) (81).

C. elegans mutant and transgenic strains used in this work are listed here. When previously published, sources of strains are referenced. Unless a reference is provided, all strains were generated in this study.

Non-transgenic C. elegans

ZG31, hif-1(ia4) V (82)

JT307, egl-9(sa307) V (44)

GR2254, moc-1(ok366) X (33)

GR2260, cdo-1(mg622) X (33)

GR2261, cdo-1(mg622) moc-1(ok366) X (33)

GR2269, suox-1(gk738847) X (33)

CB5602, vhl-1(ok161) X (39)

USD410, cysl-1(ok762) X, outcrossed 3x for this work

USD414, rhy-1(ok1402) II; suox-1(gk738847) X

USD421, egl-9(sa307) V; suox-1(gk738847) X

USD422, vhl-1(ok161) suox-1(gk738847) X

USD430, egl-9(sa307) V; cdo-1(mg622) suox-1(gk738847) X

USD431, hif-1(ia4) egl-9(sa307) V; suox-1(gk738847) X

USD432, rhy-1(ok1402) II; egl-9(sa307) V; suox-1(gk738847) X

USD433, cth-2(mg599) II; egl-9(sa307) V; suox-1(gk738847) X

USD434, egl-9(sa307) X; cysl-1(ok762) suox-1(gk738847) X

USD512, rhy-1(ok1402) II, outcrossed 4x for this work

USD706, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X

USD920, cdo-1(rae273) moc-1(ok366) X

USD921, egl-9(sa307) V; cdo-1(rae273) X

USD922, rhy-1(ok1402) II; cdo-1(rae273) X

USD937, egl-9(rae276) V; suox-1(gk738847) X

MiniMos transgenic lines

USD531, unc-119(ed3) III; raeTi1 [Pcdo-1::CDO-1::GFP unc-119(+)]

USD719, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366); raeTi14 [Pcdo-1::CDO-1(C85Y)::GFP unc-119(+)]

USD720, unc-119(ed3) III; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD730, rhy-1(ok1402) II; unc-119(ed3) III; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD733, unc-119(ed3) III; egl-9(sa307) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD739, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi1 [Pcdo-1::CDO-1::GFP unc-119(+)]

USD766, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi32 [Pcol-10::CDO-1::GFP unc-119(+)]

USD767, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi33 [Pcol-10::CDO-1::GFP unc-119(+)]

USD776, rhy-1(ok1402) II; unc-119(ed3) III; hif-1(ia4) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD777, unc-119(ed3) III; egl-9(sa307) hif-1(ia4) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD780, rhy-1(ok1402) II; unc-119(ed3) III; cysl-1(ok762) X; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD787, unc-119(ed3) III; egl-9(sa307) V; cysl-1(ok762) X; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD808, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi40 [Pcol-10::CDO-1[C85Y]::GFP unc-119(+)]

USD810, unc-119(ed3) III; cdo-1(mg622) moc-1(ok366) X; raeTi41 [Pcol-10::CDO-1[C85Y]::GFP unc-119(+)]

USD940, unc-119(ed3) III; vhl-1(ok161) X; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD1160, unc-119(ed3) III; cysl-1(ok762) X; raeTi15 [Pcdo-1::GFP unc-119(+)]

USD1161, unc-119(ed3) III; hif-1(ia4) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

EMS-derived strains

USD659, unc-119(ed3) III; egl-9(rae213) V; raeTi1 [Pcdo-1::CDO-1::GFP unc-119(+)]

USD674, unc-119(ed3) III; egl-9(rae227) V; raeTi1 [Pcdo-1::CDO-1::GFP unc-119(+)]

USD655, rhy-1(rae209) II; unc-119(ed3) III; raeTi1 [Pcdo-1::CDO-1::GFP unc-119(+)]

USD656, rhy-1(rae210) II; unc-119(ed3) III; raeTi1 [Pcdo-1::CDO-1::GFP unc-119(+)]

USD657, rhy-1(rae211) II; unc-119(ed3) III; raeTi1 [Pcdo-1::CDO-1::GFP unc-119(+)]

USD658, rhy-1(rae212) II; unc-119(ed3) III; raeTi1 [Pcdo-1::CDO-1::GFP unc-119(+)]

CRISPR/Cas9-derived strains

USD914, cdo-1(rae273) X, CDO-1::GFP

USD926, egl-9(rae276) V, EGL-9[H487A]

USD928, unc-119(ed3) III; egl-9(rae278) V; raeTi15 [Pcdo-1::GFP unc-119(+)]

MiniMos transgenesis

Cloning of original plasmid constructs was performed using isothermal/Gibson assembly (83). All MiniMos constructs were assembled in pNL43, which is derived from pCFJ909, a gift from Erik Jorgensen (Addgene plasmid #44480) (84). Details about plasmid construction are described below. MiniMos transgenic animals were generated using established protocols that rescue the unc-119(ed3) Unc phenotype (41).

To generate a construct that expressed CDO-1 under the control of its native promoter, we cloned the wild-type cdo-1 genomic locus from 1,335 base pairs upstream of the cdo-1 ATG start codon to (and including) codon 190 encoding the final CDO-1 amino acid prior to the TAA stop codon. This wild-type genomic sequence was fused in frame with a C-terminal GFP and tbb-2 3’UTR (376 bp downstream of the tbb-2 stop codon) in the pNL43 plasmid backbone. This plasmid is called pKW24 (Pcdo-1::CDO-1::GFP).

To generate pKW44, a construct encoding the active site mutant transgene Pcdo-1::CDO-1(C85Y)::GFP, we performed Q5 site-directed mutagenesis on pKW24, following manufacturer’s instructions (New England Biolabs). In pKW44, codon 85 was mutated from TGC (cysteine) to TAC (tyrosine).

To generate pKW45 (Pcdo-1::GFP), the 1,335 base pair cdo-1 promoter was amplified and fused directly to the GFP coding sequence. Both fragments were derived from pKW24, excluding the cdo-1 coding sequence and introns.

pKW49 is a construct driving cdo-1 expression from the hypodermal-specific col-10 promoter (Pcol-10::CDO-1::GFP) (53). The col-10 promoter (1,126 base pairs upstream of the col-10 start codon) was amplified and fused upstream of the cdo-1 ATG start codon in pKW24, replacing the native cdo-1 promoter. pKW53 [Pcol-10::CDO-1(C85Y)::GFP] was engineered using the same Q5-site-directed mutagenesis strategy as was described for pKW44. However, this mutagenesis used pKW49 as the template plasmid.

Chemical mutagenesis and whole genome sequencing

To define C. elegans gene activities that were necessary for the control of cdo-1 levels, we carried out a chemical mutagenesis screen for animals that accumulate CDO-1 protein. To visualize CDO-1 levels, we engineered USD531, a transgenic C. elegans strain carrying the raeTi1 [pKW24, Pcdo-1::CDO-1::GFP] transgene. USD531 transgenic C. elegans were mutagenized with ethyl methanesulfonate (EMS) using established protocols (43). F2 generation animals were manually screened, and mutant isolates were collected that displayed high accumulation of Pcdo-1::CDO-1::GFP. We demanded that mutant strains of interest were viable and fertile.

We followed established protocols to identify EMS-induced mutations in our strains of interest (85). Briefly, whole genomic DNA was prepared from C. elegans using the Gentra Puregene Tissue Kit (Qiagen) and genomic DNA libraries were prepared using the NEBNext genomic DNA library construction kit (New England Biolabs). DNA libraries were sequenced on an Illumina Hi-Seq and deep sequencing reads were analyzed using standard methods on Galaxy, a web-based platform for computational analyses (86). Briefly, sequencing reads were trimmed and aligned to the WBcel235 C. elegans reference genome (87, 88). Variations from the reference genome and the putative impact of those variations were annotated and extracted for analysis (89–91). Here we report the analysis of 6 new mutant strains (USD655, USD656, USD657, USD658, USD659, and USD674). Among these mutant strains, we found 2 unique mutations in egl-9 (USD659 and USD674) and 4 unique mutations in rhy-1 (USD655, USD656, USD657, USD658). The allele names and molecular identity of these new egl-9 and rhy-1 mutations are specified in Fig. 1C. These genes were prioritized based on the isolation of multiple independent alleles and their established functions in a common pathway, the hypoxia and H2S-sensing pathway (36–38).

Genome engineering by CRISPR/Cas9

We followed standard protocols to perform CRISPR/Cas9 genome engineering of cdo-1 and egl-9 genomic loci in C. elegans (92–95). Essential details of the CRISPR/Cas9-generated reagents in this work are described below.

We used homology-directed repair to generate cdo-1(rae273) [CDO-1::GFP]. The guide RNA was 5’-gactacagaggatctaagaa-3’ (crRNA, IDT). The GFP donor double-stranded DNA (dsDNA) was amplified from pCFJ2249 using primers that contained roughly 40bp of homology to cdo-1 flanking the desired insertion site (96). The primers used to generate the donor dsDNA were: 5’-gtacggcaagaaagttgactacagaggatctaagaataatagtactagcggtggcagtgg-3’ and 5’-agaatcaacacgttattacattgagggatatgttgtttacttgtagagctcgtccattcc-3’. Glycine 189 of CDO-1 was removed by design to eliminate the PAM site and prevent cleavage of the donor dsDNA.

The egl-9(rae276) and egl-9(rae278) [EGL-9(H487A)] alleles were also generated by homology-directed repair using the same combination of guide RNA and single-stranded oligodeoxynucleotide (ssODN) donor. The guide RNA was 5’-tgtgaagcatgtagataatc-3’ (crRNA, IDT) The ssODN donor was 5’-gcttgccatctatcctggaaatggaactcgttatgtgaaggctgtagacaatccagtaaaagatggaagatgtataaccactatttattactg-3’ (Ultramer, IDT). Successful editing resulted in altering the coding sequence of EGL-9 to encode for an alanine rather than the catalytically essential histidine at position 487. We also used synonymous mutations to introduce an AccI restriction site that is helpful for genotyping the rae276 and rae278 mutant alleles.

C. elegans growth assays

To assay developmental rates, C. elegans were synchronized at the first stage of larval development. To synchronize animals, embryos were harvested from gravid adult animals via treatment with a bleach and sodium hydroxide solution. Embryos were then incubated overnight in M9 solution causing them to hatch and arrest development at the L1 stage (97). Synchronized L1 animals were cultured on NGM seeded with BW25113 (Wild type, Moco+) or JW0764-2 (ΔmoaA753::kan, Moco-) E. coli. Animals were cultured for 48 or 72 hours (specified in the appropriate figure legends), and live animals were imaged as described below. Animal length was measured from tip of head to the end of the tail.

To determine qualitative ‘health’ of various C. elegans strains, we assayed the ability of these strains to consume all E. coli food provided on an NGM petri dish. For this experiment, dietary E. coli was produced via overnight culture in liquid LB in a 37°C shaking incubator. 200μl of this E. coli was seeded onto NGM petri dishes and allowed to dry, producing nearly identical lawns and growth environments. Then, 5 L4 animals of a strain of interest were introduced onto these NGM petri dishes seeded with OP50. For each experiment, petri dishes were monitored daily and scored when all E. coli was consumed by the population of animals. This assay is beneficial because it integrates many life-history measures (i.e. developmental rate, brood size, embryonic viability, etc.) into a single simple assay that can be scaled and applied to many C. elegans strains in parallel.

Cysteine exposure

To determine the impact of supplemental cysteine on expression of cdo-1 and animal viability, we exposed various C. elegans strains to 0, 50, 100, 250, 500, or 1,000μM supplemental cysteine. C. elegans strains were synchronously grown on NGM media supplemented with E. coli OP50 at 20°C until reaching the L4 stage of development. Live L4 animals were then collected from the petri dishes and cultured in liquid M9 media containing 4X concentrated E. coli OP50 with or without supplemental cysteine. These liquid cultures were gently rocked at 20°C overnight. Post exposure, GFP imaging was performed as described in the Microscopy section of the materials and methods. Alternatively, animals exposed overnight to 0, 100, or 1,000μM supplemental cysteine were scored for viability after being seeded onto NGM petri dishes. Animals were determined to be alive if they responded to mechanical stimulus.

Microscopy

Low magnification bright field and fluorescence images (imaging GFP simultaneously in multiple animals) were collected using a Zeiss AxioZoom V16 microscope equipped with a Hamamatsu Orca flash 4.0 digital camera using Zen software (Zeiss). For experiments with supplemental cysteine, low magnification bright field and fluorescence images were collected using a Nikon SMZ25 microscope equipped with a Hamamatsu Orca flash 4.0 digital camera using NIS-Elements software (Nikon). High magnification differential interference contrast (DIC) and GFP fluorescence images (imaging CDO-1::GFP encoded by cdo-1(rae273)) were collected using Zeiss AxioImager Z1 microscope equipped with a Zeiss AxioCam HRc digital camera using Zen software (Zeiss). All images were processed and analyzed using ImageJ software (NIH). All imaging was performed on live animals paralyzed using 20mM sodium azide. For all fluorescence images shown within the same figure panel, images were collected using the same exposure time and processed identically. To quantify GFP expression, the average pixel intensity was determined within a set transverse section immediately anterior to the developing vulva. Background pixel intensity was determined in a set region of interest distinct from the C. elegans samples and was subtracted from the sample measurements.